Yaml

---
title: "Chapter 08: Models for the Poisson Family"
output: rmarkdown::html_vignette
bibliography: bibliography.json
vignette: >
  %\VignetteIndexEntry{Chapter 08: Models for the Poisson Family}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This is a test. [@nygren2006; @item2]


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(glmbayes)
```


## 1. Introductory Discussion

Poisson regression is the canonical GLM for modeling count data.  
The Poisson likelihood is log‑concave under the log link, making it an ideal setting for envelope‑based Bayesian sampling.  
This chapter mirrors the structure used for the Binomial vignette: we begin with the classical glm() fit, then introduce the Bayesian glmb() version, discuss prior specification, and compare posterior summaries to their classical counterparts.

We use the well‑known Dobson (1990) randomized controlled trial dataset, which appears in many GLM textbooks and is included in the base R documentation for glm().  
The dataset contains counts of subjects classified by treatment group and outcome category.

## 2. Data Setup

We reproduce Dobson’s example exactly.  
The data consist of 9 observations: three outcome levels crossed with three treatment groups.

```{r dobson}
    ## Dobson (1990) Page 93: Randomized Controlled Trial :
    set.seed(333)
    counts <- c(18,17,15,20,10,20,25,13,12)
    outcome <- gl(3,1,9)
    treatment <- gl(3,3)
    print(d.AD <- data.frame(treatment, outcome, counts))
```

The table shows the observed counts for each treatment–outcome combination.

## 3. Classical Poisson Regression

The classical glm() call uses the Poisson family with the canonical log link:

```{r glm_call}

    glm.D93 <- glm(counts ~ outcome + treatment,
                   family = poisson(link = "log"))
    summary(glm.D93)
```

### 3.1 Interpretation of the Classical Output

The glm summary reports:

- **Coefficient estimates** on the log‑scale  
- **Standard errors** derived from the Fisher information  
- **Residual deviance** and **null deviance**  
- **AIC**, which penalizes model complexity

The coefficients represent multiplicative effects on the expected count.  
For example, \(\exp(\beta)\) gives the rate ratio associated with a one-unit change in a predictor.

The residual deviance (approximately 5.13) indicates a good fit relative to the null deviance (approximately 10.58), consistent with Dobson's original analysis.

## 4. Bayesian Poisson Regression with glmb()

To fit the Bayesian analogue, we must specify a prior distribution for the regression coefficients.  
The recommended workflow uses Prior_Setup(), which constructs a Zellner‑type g‑prior aligned with the model matrix.

### 4.1 Setting the Prior

```{r Prior_Setup}

    ps <- Prior_Setup(counts ~ outcome + treatment,
                      family = poisson())
    mu <- ps$mu
    V  <- ps$Sigma
```
    
The printed Prior_Setup() output shows:

- prior means (intercept from null model, effects centered at 0)  
- prior standard deviations (scaled by \( g = (1 - pwt) / pwt \))  
- the implied pseudo-sample size \( n_{\text{prior}} \)

With the default pwt = 0.01, the prior is weakly informative.

### 4.2 Calling glmb()

```{r Call_glmb}

    glmb.D93 <- glmb(counts ~ outcome + treatment,
                     family = poisson(),
                     pfamily = dNormal(mu = mu, Sigma = V))
```

This call mirrors glm() but adds the pfamily argument.

## 5. Printing and Summarizing the Bayesian Output

```{r summary_glmb}
    print(glmb.D93)
    summary(glmb.D93)
```

The summary includes:

- **Posterior means** and **posterior modes**  
- **Posterior standard deviations**  
- **Monte Carlo error estimates**  
- **Tail probabilities** relative to the prior mean  
- **Effective number of parameters (pD)**  
- **Expected residual deviance (\(\bar{D}\))**
- **DIC**, the Bayesian analogue of AIC  

### 5.1 Posterior Mean Coefficients

Posterior means are close to the classical MLEs because:

- the prior is weak (pwt = 0.01)  
- the Poisson likelihood is strongly log‑concave  
- the sample size is modest but informative  

### 5.2 Effective Number of Parameters

The pD value is typically slightly below the nominal number of coefficients, reflecting mild shrinkage from the prior.

### 5.3 DIC

DIC is approximately 56.4, very close to the classical AIC value of 56.76.  
This agreement is expected when the prior is weak and the posterior distribution is close to Gaussian.

## 6. Comparison of Classical and Bayesian Estimates

| Quantity       | Classical glm     | Bayesian glmb        |
|----------------|-------------------|------------------------|
| Point estimate | MLE               | Posterior mean         |
| Uncertainty    | SE                | Posterior SD           |
| Model fit      | Deviance, AIC     | \(\bar{D}\), \(p_D\), DIC |
| Interpretation | log-rate ratios   | same, with shrinkage   |

The Bayesian model provides richer diagnostics (tail probabilities, posterior quantiles) while preserving the familiar GLM structure.

## 7. Credible Intervals and Posterior Interpretation

Posterior percentiles (2.5%, 50%, 97.5%) provide Bayesian credible intervals for each coefficient.  
These intervals closely track the classical Wald intervals because the posterior is nearly normal.

## 8. Concluding Remarks

This vignette demonstrates that:

- Poisson GLMs with log links are ideal for envelope‑based Bayesian sampling  
- glmb() mirrors glm() while providing full posterior inference  
- Prior_Setup() offers a principled default prior  
- Posterior summaries align closely with classical results under weak priors  

In later chapters, we extend these ideas to:

- overdispersed quasi‑Poisson models  
- informative priors  
- prediction and posterior predictive checks  
- GPU‑accelerated Poisson models for large datasets


# References