---
title: "Chapter 00: nmathopencl — Package Overview"
author: "Kjell Nygren"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chapter 00: nmathopencl — Package Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## What is `nmathopencl`?

`nmathopencl` is an R package that provides GPU-accelerated versions of R's
core mathematical and statistical functions by porting R's internal `nmath`
(Mathlib) library to OpenCL. It serves two distinct but related audiences:

- **End users** who want to call distribution functions (`dnorm`, `pgamma`,
  `rbinom`, …) on vectors dispatched to a GPU, with automatic CPU fallback
  when OpenCL is unavailable.
- **Package developers** who need a reusable OpenCL math library to build
  their own custom GPU kernels without re-porting the underlying nmath
  functions themselves.

OpenCL is vendor-neutral: the same kernels run on NVIDIA, AMD, and Intel
hardware. CPU-only execution is always supported when no OpenCL stack is
present, so the package is safe to list as a dependency even in environments
that lack a GPU.

## Three-layer architecture

The package is organized in three layers, each corresponding to a set of
vignettes:

```
┌──────────────────────────────────────────────┐
│  Layer 3 — Kernels  (inst/cl/src/)           │
│  __kernel functions for the R-callable API   │
├──────────────────────────────────────────────┤
│  Layer 2 — nmath library  (inst/cl/nmath/)   │
│  Ported nmath/Rmath functions as device-side │
│  OpenCL C functions                          │
├──────────────────────────────────────────────┤
│  Layer 1 — Upstream shims                    │
│  (inst/cl/R_shims/, R_ext/, System/, …)      │
│  Type definitions, macros, and constants     │
│  that replace C headers unavailable in       │
│  OpenCL C                                    │
└──────────────────────────────────────────────┘
```

Layer 1 is the foundation: it makes the rest of the ported code compile
under OpenCL's restricted C99 dialect without modification to the nmath
sources. Layer 2 is the library: ~180 `.cl` files implementing the full
suite of Mathlib functions. Layer 3 is the API surface: thin wrapper kernels
that map a GPU work-item index to an element of an input vector and call the
appropriate Layer 2 function.

## Namespaces

The C++ side of the package is organized into three namespaces:

| Namespace | Location | Purpose |
|-----------|----------|---------|
| `openclPort` | `openclPort.h`, `opencl_kernel_runners.cpp` | Generic OpenCL infrastructure: the universal double-scalar kernel runner, error-handling utilities, device probing, and kernel loading |
| `nmathopencl` | `nmathopencl.h`, `kernel_runners.cpp`, `kernel_wrappers.cpp` | Distribution-specific kernel runners and R-facing wrappers for all nmath functions |
| `ex_glmbayes` | `ex_glmbayes_*.cpp/.h` | Self-contained example showing how a downstream package (`glmbayes`) builds custom GLM kernels on top of the two layers above |

## R-side API families

The exported R functions are organized by distribution family, mirroring the
structure of base R's `stats` package:

| R file | Functions |
|--------|-----------|
| `normal_opencl.R` | `dnorm_opencl`, `pnorm_opencl`, `qnorm_opencl`, `rnorm_opencl` |
| `gamma_opencl.R` | `dgamma_opencl`, `pgamma_opencl`, … |
| `binomial_opencl.R` | `dbinom_opencl`, `pbinom_opencl`, … |
| `poisson_opencl.R` | `dpois_opencl`, `ppois_opencl`, … |
| `beta_opencl.R` | `dbeta_opencl`, … |
| … | (and so on for all families) |
| `special_opencl.R` | `lgammafn_opencl`, `gammafn_opencl`, … |
| `math_support_opencl.R` | `fmax2_opencl`, `fmin2_opencl`, … |

Every exported function follows the same calling convention: it accepts a
scalar parameter set, dispatches to the GPU via the kernel infrastructure,
and falls back to the corresponding `stats::` or base-R function if OpenCL
is unavailable or fails.

## Checking OpenCL availability

```{r, eval = FALSE}
library(nmathopencl)

# Quick boolean check
has_opencl()

# Full diagnostic report (hardware, drivers, paths, runtime probe)
diagnose_glmbayes()
```

`has_opencl()` returns `TRUE` only when the package was compiled with
`-DUSE_OPENCL` **and** a functional OpenCL platform is detected at runtime.
All exported math functions check this internally and honor their `fallback`
argument accordingly.

## Roadmap and phased release

`nmathopencl` is the first package in a planned suite:

1. **`nmathopencl`** (this package) — released on R-universe/CRAN first.
   It temporarily hosts some generic OpenCL infrastructure (`openclPort`
   namespace) that will later migrate to a dedicated package.
2. **`glmbayes`** (already on CRAN) — will be updated to import
   `nmathopencl` for its OpenCL math dependencies, replacing its current
   bundled copy of the nmath OpenCL files.
3. **`openclPort`** (in development) — will become a standalone CRAN
   package providing the generic OpenCL porting framework. At that point,
   the `openclPort` namespace content will migrate out of `nmathopencl`.

## Vignette guide

| Vignette | Topic |
|----------|-------|
| Chapter 00 (this document) | Package overview and architecture |
| Chapter 01 | Setting up OpenCL and enabling GPU acceleration |
| Chapter 02 | Upstream shims: porting C headers to OpenCL C |
| Chapter 03 | The `nmath` OpenCL library (`inst/cl/nmath/`) |
| Chapter 04 | Kernels: writing and using `__kernel` functions |
| Chapter 05 | The `nmathopencl` R API: distribution functions on the GPU |
| Chapter 06 | Kernel loading: `load_kernel_source` and `load_kernel_library` |
| Chapter 07 | Generic OpenCL kernel runners (`openclPort` layer) |
| Chapter 08 | Case study: building custom GLM kernels (`ex_glmbayes`) |
| Chapter 12 | GPU acceleration in `glmbayes` (end-user setup guide) |
| Chapter A08 | Envelope-related functions (glmbayes example internals) |
| Chapter A10 | Accelerated EnvelopeBuild with OpenCL |
