Yaml

---
title: "Chapter 1: Getting started with glmbayes"
author: "Kjell Nygren"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chapter 1: Getting started with glmbayes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup,echo = FALSE}
library(glmbayes)
```

# 1. Introduction

The **glmb** function and related *method functions* that handle the output are designed to be Bayesian versions of the **glm** function and many of its *method functions*. This vignette shows how the basic setup/calling of the functions compare and how some of the basic output is displayed.

# 2. Preparing a dataframe

To understand how the outputs of the glmb function mirrors those for the glm function, it is useful to take a look at the first portion of the example that is provided with the glm function. The data is based on Randomized Controlled Trial data from Dobson (1990). Here is a view of the data:

```{r dobson}
## Dobson (1990) Page 93: Randomized Controlled Trial :
counts <- c(18,17,15,20,10,20,25,13,12)
outcome <- gl(3,1,9)
treatment <- gl(3,3)
print(d.AD <- data.frame(treatment, outcome, counts))
```

# 3. Calling the two functions

The example code for the glm function specifies a Poisson regression model for this data. The below code chunks shows how the glmb function can be used in much the same fashion but with some extra requirements. In particular, we need to provide a prior distribution (in this case a Multivariate Normal prior) and related constants (in this case, a prior Mean and and a prior Variance-Covariance matrix) for the coefficients of interest. Optionally, we can also tell the function how many draws to make from the posterior distribution (the default used here generates 1000 iid draws). 

## 3.1 Using the classical glm function

## 3.1.1 Calling the classical glm function
Here is the call to the classical *glm* function:

```{r glm_call,results = "hide"}
## Call to glm
glm.D93 <- glm(counts ~ outcome + treatment, 
              family = poisson())
```

## 3.2 Setting the Prior and Calling the glmb function

To use the *glmb* function, we first use a call to a function *Prior_Setup" to initialize the prior and to get the variable names needed. We use the default prior from the Prior_Setup function (discussed further in a separate vignette).

### 3.2.1 Setting the prior

```{r Prior_Setup,results = "hide"}
## Using glmb
## Step 1: Set up Default Prior 
ps=Prior_Setup(counts ~ outcome + treatment,family=poisson())
mu=ps$mu
V=ps$Sigma
```
### 3.2.2 Calling the glmb function

We now call the *glmb* function and include the default prior
distribution in addition to the two required arguments for the *glm* function.

```{r Call_glmb,results = "hide"}
# Step 3: Call the glmb function
glmb.D93<-glmb(counts ~ outcome + treatment, family=poisson(), pfamily=dNormal(mu=mu,Sigma=V))
```


In the above, it is worth noting the additional step that we went through when specifying the prior. We used a call to the function *Prior_Setup* to get the correct dimensions for the mean and variance-covariance matrices and to initialize the constants. The *Prior_Setup* function also provided information on the Variable names in the design matrix (which also corresponds to the names of the coefficients eventually estimated) so we can make informed changes to the prior if so desired.

The next couple of vignette's will discuss the prior specification in more details. Here we instead focus on reviewing the output from the function and how it can be used.

# 4. Printing the output

Taking a look at the basic printed output, we can see that the two closely mirror each other with the glmb posterior means replacing the glm maximum likelihood estimates.


## 4.1 Printed glm output

```{r Printed_glm_Views}
## Printed view of the output from the glm function 
print(glm.D93)
```
### 4.1.1 Coefficients

### 4.1.2 Null and Residual Degrees of Freedom

### 4.1.3 Residual Deviance

### 4.1.4 AIC


## 4.2 Printed glmb output

```{r Printed_glmb_Views}

## Printed view of the output from the glmb function 
print(glmb.D93)
```
### 4.2.1 Posterior Mean Coefficients

### 4.2.2 Effective Number of Parameters

### 4.2.3 Expected Residual Deviance

### 4.2.4 DIC


In addition to the posterior means, the *glmb* printed output also returned three pieces of information that are similar to (but not quite the same) as the classical output. The "Effective Number of Parameters" should in general be close to (but not exactly the same) as the number of parameters estimated while the Expected Residual Deviance in general will be higher than the corresponding maximum likelihood estimate for the Residual Deviance (since the latter is designed to minimize it). Finally, the DIC is essentially a Bayesian version of the AIC. These measures will be discussed in greater detail in one of our later vignettes.


# 5. Summarizing the output
Let's take a closer look at the outputs of the summary functions.


## 5.1 Summary view of the glm output

In turn, we see the Call, a list of Deviance residuals, and the estimated coefficients. The coefficients are then followed by some additional model related information. 

```{r glm_summary}
## summary output for the "glm" class
summary(glm.D93)
```

## 5.2 Summary view of the glmb output

The summary for the glmb function follow a similar structure but adds a table containing information related to the prior and the maximum likelihood above the table with the means for the estimated Bayesian coefficients. The output below the main table with coefficients is also modified to contain similar (but slightly different) pieces of information (the details of which are discussed elsewhere).

```{r glmb_summary}
## summary output for the "glm" class
summary(glmb.D93)
```

# 6. Textbook Content Mappings

## 6.1 Agresti - Foundations of Linear and Generalized Linear Models

| Vignette Section                           | Textbook Chapter & Section                                        | Notes                    |
|--------------------------------------------|-------------------------------------------------------------------|--------------------------|
| 3.1.1 Calling the glm function             | Section 8.1 Poisson Log-Linear Models| AAA                           |
| 3.2.1 Setting the prior                    | Section 11.2 Prior Distributions for Regression Coefficients   | See Agresti’s discussion of g-priors.|
| 3.2.2 Calling the glmb function            | Section 11.3: Bayesian Generalized Linear Models| AAA  |
| 4.1.1 Coefficients                         | Section 8.4: Parameter Intepretation in Count Models | AAA                  |
| 4.1.2 Null and Residual Degrees of Freedom | Section 5.2: Deviance and Likelihood Ratio Tests | AAA                  |
| 4.1.3 Residual Deviance                    | Section 7.5: Deviance and Goodness of Fit for Count Data | AAA                      |
| 4.1.4 AIC                                  | Section 5.6: Variable Selection and Penalized Methods | AAA                           |
| 4.2.1 Posterior Mean Coefficients          | Section 11.2: Bayesian Linear Models | AAA                  |
| 4.2.2 Effective Number of Parameters       | Section 11.3: Bayesian Generalized Linear Models | AAA                  |
| 4.2.3 Expected Residual Deviance           | Section 11.3: Bayesian Generalized Linear Models  | AAA                      |
| 4.2.4 DIC                                  | Section 11.3: Bayesian Generalized Linear Models  | AAA                           |


# 7. References
