---
title: "Chapter 01: Setting Up OpenCL and Enabling GPU Acceleration"
author: "Kjell Nygren"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chapter 01: Setting Up OpenCL and Enabling GPU Acceleration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

`nmathopencl` can run in two modes:

- **CPU-only mode**: all distribution functions fall back to `stats::` or
  base-R equivalents. No special setup is needed. This is the default when
  the package is installed from a prebuilt binary (e.g., from CRAN or
  R-universe).
- **GPU-accelerated mode**: kernels execute on an OpenCL device. This
  requires compiling the package from source on a machine with an OpenCL
  SDK installed.

This vignette describes how to satisfy the prerequisites and verify a
working GPU setup.

## What "compiling with OpenCL" means

At compile time, the package's `configure` script detects whether
`CL/cl.h` is present and a linkable `OpenCL` library exists. If both are
found, it sets `-DUSE_OPENCL` and `-DGLMBAYES_BUILD` in the compile flags
and links against `-lOpenCL`. All OpenCL-specific C++ code is guarded by
`#ifdef USE_OPENCL`, so the package compiles cleanly without OpenCL headers.

A binary built **without** `USE_OPENCL` is a fully valid CPU-only package.
`has_opencl()` returns `FALSE` and every math function silently uses its
`stats::` fallback.

## Prerequisites

Enabling GPU acceleration requires three independent pieces:

| Component | What it provides | Where to get it |
|-----------|-----------------|-----------------|
| **GPU driver** | Exposes the hardware to the OS | Vendor website or OS package manager |
| **OpenCL headers** (`CL/cl.h`) | Needed at compile time | Part of the GPU vendor SDK or `opencl-headers` package |
| **OpenCL ICD loader / runtime** | Needed at runtime (`OpenCL.dll` / `libOpenCL.so`) | Installed with driver or separately as `ocl-icd-libopencl1` |

All three must be present. Having the headers but not the loader (or vice
versa) is the most common failure mode.

## Platform-specific instructions

### Windows (NVIDIA)

1. Install the **NVIDIA Studio Driver** (recommended for stability) or
   Game Ready Driver from <https://www.nvidia.com/Download>.
2. Install the **CUDA Toolkit** from <https://developer.nvidia.com/cuda-downloads>.
   The toolkit ships `CL/cl.h` under `<cuda_root>/include/` and provides
   the ICD loader.
3. Ensure `<cuda_root>/include` and `<cuda_root>/lib/x64` are visible to
   the compiler and linker. The `configure.win` script searches common
   CUDA Toolkit paths automatically.

### Windows (AMD)

1. Install the AMD driver from <https://www.amd.com/support>.
2. Install the **AMD APP SDK** or the
   [Khronos OpenCL SDK](https://github.com/KhronosGroup/OpenCL-SDK/releases)
   to obtain `CL/cl.h`.
3. `OpenCL.dll` is installed with the AMD driver into `System32`.

### Windows (Intel)

1. Install the Intel graphics driver from <https://www.intel.com/content/www/us/en/download-center/home.html>.
2. Install the [Intel oneAPI Base Toolkit](https://www.intel.com/content/www/us/en/developer/tools/oneapi/base-toolkit.html)
   for headers and runtime.

### Linux (NVIDIA)

```bash
# Install driver
sudo ubuntu-drivers autoinstall        # Ubuntu
# or
sudo apt install nvidia-driver-550

# Install OpenCL headers and ICD loader
sudo apt install opencl-headers ocl-icd-libopencl1 ocl-icd-dev

# The NVIDIA OpenCL ICD is shipped with the driver
# Verify with:
ls /etc/OpenCL/vendors/nvidia*.icd
```

### Linux (AMD)

```bash
sudo apt install opencl-headers ocl-icd-libopencl1
# AMD ROCm provides the vendor ICD:
# Follow https://rocm.docs.amd.com/en/latest/
```

### Linux (Intel)

```bash
sudo apt install opencl-headers ocl-icd-libopencl1 intel-opencl-icd
```

### WSL2 (Windows Subsystem for Linux)

GPU compute in WSL2 is supported for **NVIDIA only**. The NVIDIA driver
must be installed in Windows (not inside WSL). The WSL-specific driver stub
(`/usr/lib/wsl/lib/libcuda.so`) provides OpenCL access.

```bash
# Inside WSL2:
sudo apt install opencl-headers ocl-icd-libopencl1
# The NVIDIA ICD is provided by the Windows host driver via /usr/lib/wsl/lib/
```

AMD and Intel GPUs cannot be exposed to WSL2.

## Installing `nmathopencl` from source

Once the OpenCL SDK is installed, install the package from source:

```{r, eval = FALSE}
# From a local clone:
devtools::install("C:/Rpackages/nmathopencl")

# Or from R-universe (when available):
install.packages("nmathopencl",
                 repos = "https://<username>.r-universe.dev")
```

The `configure` / `configure.win` script runs automatically and will print
a message indicating whether `USE_OPENCL` was activated.

## Verifying the setup

```{r, eval = FALSE}
library(nmathopencl)

# 1. Quick check
has_opencl()
#> [1] TRUE  (if compiled with USE_OPENCL and a device is available)

# 2. List detected GPU devices
gpu_names()
#> [1] "NVIDIA GeForce RTX 4090"

# 3. Full diagnostic report
diagnose_glmbayes()
```

`diagnose_glmbayes()` runs a layered check:

1. Detects the OS environment (Windows / Linux / WSL2 / MSYS2).
2. Identifies GPU vendors (NVIDIA / AMD / Intel) via `nvidia-smi`,
   `lspci`, or WMI (`Win32_VideoController`).
3. Checks whether the GPU driver is installed.
4. Verifies that `CL/cl.h` (compile-time header) is present.
5. Verifies that the OpenCL ICD loader / runtime DLL is present.
6. On Linux/WSL2, compiles and runs a minimal OpenCL probe program to
   confirm that a platform is actually callable.
7. Checks PATH and `LD_LIBRARY_PATH` for required directories.
8. Reports the compile-time `has_opencl()` status.

On Windows, step 6 is skipped because platform-creation probes are less
reliable there; rely on the driver and ICD checks instead.

## Troubleshooting common failures

### `has_opencl()` returns `FALSE` after driver installation

The most likely cause is that the package was compiled before the SDK was
installed, or from a prebuilt binary. Reinstall from source:

```{r, eval = FALSE}
install.packages("nmathopencl", type = "source")
```

### Compilation fails with `CL/cl.h: No such file or directory`

The OpenCL headers are not on the compiler include path. Install the SDK
(see above) and, if necessary, set the `OPENCL_HOME` or `OPENCL_SDK`
environment variable to the SDK root before installing.

### Runtime error: `clGetPlatformIDs: CL_PLATFORM_NOT_FOUND_KHR`

Headers were found at compile time but no runtime platform is available.
Check:
- Is the GPU driver installed and up to date?
- Is the ICD file present (e.g., `/etc/OpenCL/vendors/nvidia.icd` on Linux)?
- On WSL2: is the NVIDIA driver installed in Windows, not just WSL?

### `PATH` warnings from `diagnose_glmbayes()`

On Windows, the CUDA Toolkit bin directory (e.g.,
`C:/Program Files/NVIDIA GPU Computing Toolkit/CUDA/v12.9/bin`) must be in
`PATH` so that `OpenCL.dll` is found at runtime. `diagnose_glmbayes()` will
offer to add missing entries permanently.

## After a successful setup

Once `has_opencl()` returns `TRUE`, all `*_opencl` functions automatically
dispatch to the GPU:

```{r, eval = FALSE}
x <- rnorm(1e6)
system.time(dnorm_opencl(x, mean = 0, sd = 1))
system.time(dnorm(x, mean = 0, sd = 1))
```

For large vectors the GPU path is substantially faster. For small vectors
the overhead of kernel compilation and data transfer may make it slower;
the `fallback = TRUE` argument (the default) ensures graceful degradation.
