Yaml

---
title: "Chapter A1: Overview of Estimation Procedures"
author: "Kjell Nygren"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chapter A1: Overview of Estimation Procedures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(glmbayes)
```


# Chapter A1: Simulation Functions for Bayesian Generalized Linear Models

Kjell Nygren  
2026-01-XX

This chapter provides an overview of the simulation functions ("simfunctions") that form the computational
core of the glmbayes package. These functions generate independent and identically distributed (iid) samples
from posterior distributions for Bayesian generalized linear models (GLMs). They are lower-level, high-performance
building blocks used internally by the higher-level modeling functions `glmb()` and `lmb()`, and they are
also suitable for use in Gibbs samplers, block samplers, and other custom simulation-based inference methods.

The simfunctions implement both classical conjugate sampling procedures and the envelope-based accept-reject
methods developed in Nygren and Nygren (2006). The envelope-based methods are documented in detail in
Appendix A5 (coefficient samplers), Appendix A6 (Gamma-family dispersion sampler), and Appendix A7
(joint sampling for Gaussian regression with independent Normal-Gamma priors).

All functions described here are implemented in the file `simfunctions.R`, and many call optimized C or C++
routines for envelope construction and sampling. These routines include:

- `EnvelopeBuild` and `EnvelopeBuild_c.cpp`
- `EnvelopeDispersionBuild_cpp`
- `.rnnorm_reg_cpp`
- `.rindep_norm_gamma_reg_std_cpp`
- `.rindep_norm_gamma_reg_std_parallel_cpp`
- `.rnorm_reg_cpp`
- The numerically robust truncated-Gamma sampler `ctrgamma`

The purpose of this chapter is to describe how these functions fit together, how they map to families and
pfamilies, and how they relate to the theoretical results in the JASA paper and the appendix chapters.


## 1. What the Simulation Functions Do

The simulation functions provide a unified interface for generating posterior samples for:

1. Regression coefficients beta
2. Dispersion or precision parameters phi or tau
3. Joint sampling of (beta, phi) when required by the prior structure

They are designed to:

- Produce iid samples (not MCMC chains)
- Use conjugate sampling when available
- Use envelope-based accept-reject sampling when conjugacy is not available
- Provide minimal overhead for use inside Gibbs samplers or custom algorithms
- Expose internal quantities such as posterior modes, envelopes, iteration counts, and standardized models

The higher-level functions `glmb()` and `lmb()` wrap these simfunctions with additional overhead:
model frames, formulas, fitted objects, diagnostics, and S3 methods. The simfunctions avoid this overhead
and are therefore preferred when building custom samplers or when performance is critical.


## 2. Mapping Families and Pfamilies to Simulation Functions

The table below summarizes how likelihood families, link functions, and pfamilies map to the
simulation functions in `simfunctions.R`. It also indicates whether the sampling is conjugate or
envelope-based, and lists the C or C++ routines used when applicable.

+----------------------+----------------------+------------------------------+------------------------------+---------------------------+-------------------------------------------+
| Likelihood Family    | Link                 | Pfamily                      | Simfunction                  | Method                    | Backend R / C++ Routines                  |
+----------------------+----------------------+------------------------------+------------------------------+---------------------------+-------------------------------------------+
| Gaussian             | identity             | dNormal                      | rNormal_reg                  | Conjugate MVN             | .rnorm_reg_cpp                            |
| Gaussian             | identity             | dNormal_Gamma                | rNormal_Gamma_reg            | Conjugate Normal-Gamma    | Pure R                                    |
| Gaussian             | identity             | dIndependent_Normal_Gamma    | rindependent_norm_gamma_reg  | Envelope-based joint       | .rindep_norm_gamma_reg_std_cpp            |
|                      |                      |                              |                              | sampler for (beta,phi)    | EnvelopeDispersionBuild_cpp               |
+----------------------+----------------------+------------------------------+------------------------------+---------------------------+-------------------------------------------+
| Poisson              | log                  | dNormal                      | rNormal_reg                  | Envelope-based             | .rnnorm_reg_cpp, EnvelopeBuild            |
| Quasi-Poisson        | log                  | dNormal                      | rNormal_reg                  | Envelope-based             | .rnnorm_reg_cpp, EnvelopeBuild            |
+----------------------+----------------------+------------------------------+------------------------------+---------------------------+-------------------------------------------+
| Binomial             | logit, probit,       | dNormal                      | rNormal_reg                  | Envelope-based             | .rnnorm_reg_cpp, EnvelopeBuild            |
|                      | cloglog              |                              |                              |                           |                                           |
+----------------------+----------------------+------------------------------+------------------------------+---------------------------+-------------------------------------------+
| Gamma                | log                  | dNormal                      | rNormal_reg                  | Envelope-based             | .rnnorm_reg_cpp, EnvelopeBuild            |
| Gamma                | log                  | dGamma                       | rGamma_reg                   | Conjugate or A6 envelope  | ctrgamma, A6 routines                     |
+----------------------+----------------------+------------------------------+------------------------------+---------------------------+-------------------------------------------+

This table provides the "at a glance" map for determining which simulation function is used for a
given model. Conjugate cases use direct sampling from known posterior distributions. All other
cases rely on the likelihood-subgradient envelope methods described in Nygren and Nygren (2006)
and in Appendices A5, A6, and A7.

## 3. Conjugate vs. Envelope-Based Sampling

### 3.1 Conjugate Cases

Several combinations of likelihood family and prior lead to closed-form posterior distributions:

- Gaussian likelihood with Normal prior on beta
- Gaussian likelihood with Normal-Gamma prior on (beta, tau)
- Gaussian likelihood with Gamma prior on tau
- Gamma likelihood with Gamma prior on tau (conjugate path)

In these cases, the simfunctions use direct sampling from multivariate normal or gamma distributions.
No envelope construction is required.

### 3.2 Envelope-Based Cases

When the posterior is not available in closed form, the simfunctions use the likelihood subgradient
approach of Nygren and Nygren (2006). These cases include:

- All non-Gaussian GLMs with Normal priors
- Gaussian regression with independent Normal-Gamma priors
- Gamma regression with Gamma prior on precision (non-conjugate path)

The envelope-based samplers rely on:

- Likelihood subgradient densities (JASA 2006)
- Mixture envelopes (Appendix A5)
- Dispersion envelopes (Appendix A6)
- Joint envelopes for (beta,phi) (Appendix A7)

These methods guarantee iid sampling even in high dimensions and for non-conjugate models.


## 4. Theoretical Foundations and Cross-References

The simulation functions are grounded in the following theoretical results:

- Nygren and Nygren (2006), "Likelihood Subgradient Densities", Journal of the American Statistical Association.
  This paper introduces the likelihood-subgradient approach and proves bounds on acceptance rates.

- Appendix A5: Technical report for likelihood-subgradient based samplers.
  Details the envelope construction for regression coefficients.

- Appendix A6: Accept-reject sampling for dispersion in Gamma regression.
  Provides the theory behind the dispersion sampler used in `rGamma_reg`.

- Appendix A7: Accept-reject sampling for Gaussian regression with independent Normal-Gamma priors.
  Provides the theory behind the joint sampler used in `rindependent_norm_gamma_reg`.

These references provide the mathematical justification for the envelope-based methods implemented in the simfunctions.


## 5. Overview of Each Simulation Function

### 5.1 rNormal_reg

**Purpose:**  
Generates iid samples of regression coefficients beta for GLMs with Normal priors.

**Conjugate case:**  
Gaussian likelihood with Normal prior.  
Sampling uses direct multivariate normal draws via `.rnorm_reg_cpp`.

**Envelope-based case:**  
All non-Gaussian GLMs.  
Uses `.rnnorm_reg_cpp` and `EnvelopeBuild` to construct a likelihood-subgradient envelope.

**Implementation sketch:**  
1. Compute posterior mode using `optim` (non-Gaussian) or `lm.fit` (Gaussian).  
2. Standardize model using `glmb_Standardize_Model`.  
3. Build envelope using `EnvelopeBuild`.  
4. Generate iid samples using accept-reject sampling.

**References:**  
JASA 2006, Appendix A5.


### 5.2 rNormal_Gamma_reg

**Purpose:**  
Conjugate Normal-Gamma regression for Gaussian likelihoods.

**Method:**  
Closed-form posterior for (beta, tau).  
Draw tau from Gamma, then draw beta from conditional Normal.

**Implementation sketch:**  
1. Fit weighted least squares to obtain Btilde, IR, and S.  
2. Compute posterior Gamma parameters for tau.  
3. Draw tau and then beta = Btilde + IR %*% rnorm * sqrt(dispersion).

**References:**  
Classical Normal-Gamma theory.


### 5.3 rindependent_norm_gamma_reg

**Purpose:**  
Joint sampling of (beta, phi) for Gaussian regression with independent Normal-Gamma priors.

**Method:**  
Envelope-based joint sampler using a shared envelope over beta and dispersion bounds.

**Backend:**  
- `.rindep_norm_gamma_reg_std_cpp`  
- `EnvelopeDispersionBuild_cpp`  
- `EnvelopeBuild`

**Implementation sketch:**  
1. Iteratively anchor dispersion using repeated Gaussian fits.  
2. Standardize model and compute posterior mode.  
3. Build coefficient envelope (EnvelopeBuild).  
4. Build dispersion envelope (EnvelopeDispersionBuild_cpp).  
5. Sample jointly from the standardized model.  
6. Transform back to original scale.

**References:**  
Appendix A7.


### 5.4 rGamma_reg

**Purpose:**  
Simulates dispersion parameters for Gaussian and Gamma families.

**Gaussian case:**  
Conjugate Gamma posterior for precision tau.  
Uses direct sampling or truncated sampling via `ctrgamma`.

**Gamma family case:**  
Non-conjugate.  
Uses the A6 envelope-based sampler with a tangent envelope for the precision.

**Backend:**  
- `ctrgamma`  
- A6 routines  
- log-space utilities `logdiffexp`

**Implementation sketch:**  
1. Compute posterior mode for precision.  
2. Build Gamma surrogate and tangency point.  
3. Construct bounding function log_h.  
4. Accept-reject sampling for precision.  
5. Transform to dispersion.

**References:**  
Appendix A6.


## 6. Use in Gibbs Samplers and Custom Algorithms

The simfunctions are designed to be lightweight and fast. They avoid the overhead of:

- Model frames
- Formulas
- S3 objects
- Diagnostic structures

This makes them ideal for:

- Gibbs sampling
- Blocked MCMC
- Variational approximations
- Simulation studies
- Algorithm prototyping

They return only the essential components needed for further computation:
coefficients, dispersion draws, posterior modes, envelopes, and iteration counts.


## 7. Source Code and File Structure

The simfunctions are implemented in:

- `simfunctions.R`

They call supporting routines in:

- `rnnorm_reg_cpp.cpp`
- `rnorm_reg_cpp.cpp`
- `rindep_norm_gamma_reg_std_cpp.cpp`
- `EnvelopeBuild_c.cpp`
- `EnvelopeDispersionBuild_cpp.cpp`
- `glmb_Standardize_Model.R`

Utility functions include:

- `ctrgamma` (truncated Gamma sampler)
- `logdiffexp` (stable log difference)
- `p_inv_gamma`, `q_inv_gamma`, `r_invgamma` (inverse-Gamma utilities)


## 8. Summary

The simulation functions described in this chapter form the computational backbone of the glmbayes package.
They implement both classical conjugate sampling and modern envelope-based iid sampling. They are the
recommended interface for custom samplers, Gibbs sampling, and research applications requiring direct access
to the underlying algorithms.

For further details, see the JASA paper and Appendices A5, A6, and A7.





## General Discussion

The estimation procedures implemented in this package can divided into the following four distinct categories:

* Finding posterior modes
* Simulating from models with with conjugate priors
* Generating iid samples for log-concave models with non-conjugate Normal, Gamma, or Independent_Normal_Gamma priors 
* Bounding Distances to posterior densities for two-block Gibbs samplers

In this vingette, we provide a brief overview of how these procedures are implemented. It is worth noting that the 
combination of these four approaches likely can be applied to wider range of models and future plans for this package include the implementation of additional models.

## Finding Posterior Modes

Finding posterior modes is an important first component for most of the implemented sampling procedures in this package.   The posterior modes are returned as part of the output itself from many of the core functions and are also used to center many of the sampling procedures implemented so they also help determine the accuracy and speed of underlying sampling procedures.  The process for fiding posterior modes work a bit differently depending on the *family* and *pfamily* in question. Our subsections here describe the currently implemented approaches

### Posterior Modes for the *gaussian()* family combined with either the *dNormal()* or *dNormal_Gamma()* pfamilies

In this case, the posterior modes have closed form solutions that also correspond to the posterior means for the posterior distributions. Moreover, using proper priors ensure that the modes always exists (unlike the classical case where unique maximum likelihood estimates might not exist).  The concern in this case is mainly with ensuring numerical accuracy, especially in cases where the model might have many variables or where the resulting model coefficients (despite the prior) might be poorly identified. 

Our implemented process for finding posterior modes essentially involves the following steps

* Perform a Cholesky decomposition for the prior precision (inverse variance-covariance) 
* Use the Cholesky decomposition to add additional "prior" observations to the dataset
* Use the *lm.fit* or *lm.wfit* functions for the modified dataset in order to return resulting posterior mode estimates

The source codes *rnorm_reg_cpp.cpp* (for the *dNormal()* prior)  and *rNormal_reg.wfit.R* (for the *dNormal_Gamma()* Prior) shows the current version of these implementations. Note that the *rNormal_reg.wfit* function is called from within the *rNormal_Gamma_reg* function in order to find the posterior mode.

In addition to having the benefit of the robust estimation procedures built into the *lm.fit* and *lm.wfit* functions, calling these functions also have the added benefit of returning the components needed to have access to the various influence related methods available for classes *lm* and *glm*.

### Posterior Modes for *non-gaussian* families combined with the *dNormal()* pfamily.

In this case, there is no closed form solution for the posterior mode. Our current implementation (see the source code *rnnorm_reg_cpp.cpp*) leverages a call to the *optim* function with the *method="BFGS"* and *hessian=TRUE* options in order to attempt to find the posterior mode. The *"BFGS"* method [Add reference] is a quasi-Newton method (also known as a variable metric algorithm) that uses function values and gradients to build up a picture of the surface optimized. The *hessian=TRUE* option returns an approximation for the hessian associated with  the negative of the log-posterior density which is leveraged during standardization of the model prior to the simuation phase (see the section "Generating iid Samples for Non-Conjugate Families" below for details) for these models.

While the posterior mode is derived using the above method (and used as the estimate for the posterior mode), our *rNormal_reg* function also uses a call to the *glmb.wfit* function in order to get the output needed for the influence measures. This implements a similar procedure to that in the *glm.fit* function before doing a single call to the *lm.fit* function. Similar to the procedure in the *gaussian()* family case above, the data provided to the *lm.fit* function has additional added observations (based on a Cholesky decomposition). We refer the reader to the source codes for the *glm.fit* and *glmb.wfit* for the details on the similarities and differences between the two functions. Essentially, the latter assumes that the posterior mode has alredy been reached (using the calls to the optim function) so no iterations are therefore required.

### Conditional Posterior Modes for the *gaussian* family combined with the *dIndependent_Normal_Gamma()* pfamily.

The rIndependent_Normal_Gamma_reg procedure currently uses an initial iterative process where the coefficients and dispersion are conditionally optimized in order to arrive at a good centering of candidate densities for the simulation. The conditional optimization of the coefficients in this process relies on the same process as in the other *gaussian()* family case above. The shape and rate parameters for the candidate gamma distribution are in turn set based on the expected *RSS* based on simulation from the conditional densitiy associated with the conditionally optimized coefficients (see the current source code *rIndependent_Normal_Gamma_Reg.R* for details). 

Plese note that the methods implemented for this prior should be regarded as still being in the developemental/experimental phase. The current implementation appears to yield valid *iid* samples at reasonable speeds for some of our smaller models, but it is unclear how well this will perform for larger models and methods may be adjusted in the future to improve performance.

## Simulating from Models With Conjugate priors

Three of the currently implemented combinations of *families* and *pfamilies* involves what is known in the literature as conjugate prior distributions where the posterior distributions take on the same well known probability distribution as the prior distributions [ADD REFERENCES]. These models are

* The *gaussian()* family with a *dNormal()* pfamily prior for the regression coefficients
* The *gaussian()* family with a *dGamma()* pfamily prior for the inverse dispersion (precision) parameter
* The *gaussian()* family with a *dNormal_Gamma()* pfamily prior the regression coefficienst and inverse dispersion parameter.

Simulation in these cases reduces to simulation from *multivariate normal* and/or *gamma* distributions and the main concern is ensuring numerical accuracy and speed for the resulting simulation. We refer the reader to the source codes 
*rnorm_reg_cpp.cpp*, *rgamma_reg.R*, and *rNormal_Gamma_reg* for details on the current implementation of these simulation procedures in the current package.

## Generating iid Samples for Non-Conjugate Families

The most innovative components of the current package likely involves the procedures implemented in order to generate *iid* samples from models with various  non-gaussian families combined with *dNormal()* pfamily priors as well as for models combining the *gaussian()* family with *dIndependent_Normal_Gamma()* pfamily prior distributions. In addititon the package contains an *iid* sampling procedure for the *Gamma()* family with a *dGamma()* pfamily prior for the dispersion parameter. We discuss each of these at a high level here and provide references to additional details for those interested.

### Non-gaussian families with Normal Priors

 When the specified *family* is any log-concave non-gaussian family, then the estimation uses the Likelihood
 subgradient density approach of Nygren and Nygren (2006).  This approaches uses tangencies to the 
 log-likelihood function in order to construct an enveloping function from which candidate draws are 
 generated and then either accepted/rejected using accept/reject methods. The core C function performing
 this simulation essentially goes through the following steps: 
  
 
 1) The model is standardized to have prior mean vector equal to 0 (i.e., offsets and any prior mean are combined
 into a constant term).
 
 
 2) The posterior mode for this transformed model is found. Currently this uses the *optim* function. 
 Later implementations may replace this with iteratively reweighted least squares (IWLS) to increase 
 consistency with the *glm* function and to enhance numerical accuracy.
 
 
 3) The model is further standardized so that (a) the precision matrix at the posterior mode is diagonal and (b)
 the prior variance-covariance matrix is the identity matrix (see the documentation *glmb_Standardize_Model* and the discussion below for details).
 
 
 4) An enveloping function is built for the the standardized model, containing constants needed during simulation (see
 the documentation *EnvelopeBuild* and the discussion below for details).
 
 
 5) Samples for the standardized model are generated using accept-reject methods (see the documentation for *rnnorm_reg_std* and the discussion below for details).
 
 
 6) The output from the standardized model are transformed back to the original scale by reversing the two 
 eigenvalue decompositions and by adding back the prior mean.  


The below code chunks provide a more detailed illustration of the estimation process

**Setup of Data and Prior** 
```{r Setup data and prior}
data(menarche,package="MASS")

Age2=menarche$Age-13
x<-matrix(as.numeric(1.0),nrow=length(Age2),ncol=2)
x[,2]=Age2

y=menarche$Menarche/menarche$Total
wt=menarche$Total

mu<-matrix(as.numeric(0.0),nrow=2,ncol=1)
mu[2,1]=(log(0.9/0.1)-log(0.5/0.5))/3
V1<-1*diag(as.numeric(2.0))

# 2 standard deviations for prior estimate at age 13 between 0.1 and 0.9
## Specifies uncertainty around the point estimates

V1[1,1]<-((log(0.9/0.1)-log(0.5/0.5))/2)^2 
V1[2,2]=(3*mu[2,1]/2)^2  # Allows slope to be up to 1 times as large as point estimate 

dispersion2<-as.numeric(1.0)
offset2=rep(as.numeric(0.0),length(y))
P=solve(V1)
n=1000


```

**Gathering of family functions Used in process** 

```{r Setup_Family_Functions}

famfunc<-glmbfamfunc(binomial(logit))

f1<-famfunc$f1
f2<-famfunc$f2  # Used in optim and glmbsim_cpp
f3<-famfunc$f3  # Used in optim
f5<-famfunc$f5
f6<-famfunc$f6

```

**Finding the posterior mode** 

```{r Find Posterior Mode}
start <- mu

###### Adjust weight for dispersion

wt2=wt/dispersion2


######################### Shift mean vector to offset so that adjusted model has 0 mean

alpha=x%*%as.vector(mu)+offset2
mu2=0*as.vector(mu)
P2=P
x2=x


#####  Optimization step to find posterior mode and associated Precision

parin=start-mu

opt_out=optim(parin,f2,f3,y=as.vector(y),x=as.matrix(x),mu=as.vector(mu2),
              P=as.matrix(P),alpha=as.vector(alpha),wt=as.vector(wt2),
              method="BFGS",hessian=TRUE
)

bstar=opt_out$par  ## Posterior mode for adjusted model
bstar # Mode from Optimization
bstar+as.vector(mu)  # Mode for actual model
A1=opt_out$hessian # Approximate Precision at mode
A1

```
**Standardizing the Model** 

 This functions starts with basic information about the model in the argument list and then
uses the following steps to further standardize the model (the model is already assumed to have a 0 prior mean vector
 when this step is applied).
 
 1) An eigenvalue composition is applied to the posterior precision matrix, and the model is (as an interim step)
 standardized to have a posterior precision matrix equal to the identity matrix. Please note that this means
 that the prior precision matrix after this step is *"smaller"* than the identity matrix.
 
 
 2) A diagonal matrix epsilon is pulled out from the standardized prior precision matrix so that the remaining
 part of the prior precision matrix still is positive definite. That part is then treated as part of the posterior
 for the rest of the standardization and simulation and only the part connected to epsilon is treated as part of the prior. 
 Note that the exact epsilon chosen seems not to matter. Hence there are many possible ways of doing this 
 standardization and future versions of this package may tweak the current approach 
 if it helps improve numerical accuracy or acceptance rates.
 
 
 3) The model is next standardized (using a second eigenvalue decomposition) so that the prior (i.e., the portion connected to epsilon) is the identity  matrix. The standardized model then simutaneously has the feature that the prior precision matrix is the  identity matrix and that the data precision A (at the posterior mode) is a diagonal matrix. Hence the variables in the standardized model are approximately independent at the posterior mode.


```{r Standardize Model}
## Standardize Model

Standard_Mod=glmb_Standardize_Model(y=as.vector(y), x=as.matrix(x),P=as.matrix(P),
                                    bstar=as.matrix(bstar,ncol=1), A1=as.matrix(A1))

bstar2=Standard_Mod$bstar2  
A=Standard_Mod$A
x2=Standard_Mod$x2
mu2=Standard_Mod$mu2
P2=Standard_Mod$P2
L2Inv=Standard_Mod$L2Inv
L3Inv=Standard_Mod$L3Inv

```

**Building the Envelope** 

 To construct an enveloping function, we follow the approach in Nygren and Nygren (2006)
 which involves the following steps when a maximally sized grid is constructed 
 (if the prior for some dimensions is relatively strong, this may not be needed)
 

 1) For each dimension, a constant *omega_i* is found that depends on the 
 corresponding diagonal element in the precision matrix.
 
  
 2) Corresponding intervals *(thetastar_i-0.5 omega,thetastar_i-0.5 omega)*
  are constructed around the posterior mode thetastar for each dimension
  
  
 3) The mode as well as the points *thetastar_i-omega_i* and *thetastar_i+omega_i*
 are selected as the components of the points at which tangencies will be found for each of 
 the dimensions.  
 
 
 4) A Grid is constructed with all possible combinations of points and 
 negative log-likelihood and gradient for the negative log-likelihood are evaluated (see the
 EnvelopeBuild_c.cpp function source code for details)
 
 
 5) The *Set_Grid* function is called in order to evaluate the log of the density associated
 with each of the resulting restricted multivariate normals by evaluating the differences between the cummulative 
 density for each dimension between its lower and upper bound.
 
 
 6)  The Set_LogP function is called in order to help set the probabilities with which each of the components
 of the grid should be sampled (see remark 6 in Nygren and Nygren (2006)).
 
 
 Any constants needed by the sampling are added to a list and returned by the function. 


```{r Build Envelope}

Env2=EnvelopeBuild(as.vector(bstar2), as.matrix(A),y, as.matrix(x2),
                   as.matrix(mu2,ncol=1),as.matrix(P2),as.vector(alpha),as.vector(wt2),
                   family="binomial",link="logit",Gridtype=as.integer(3), 
                   n=as.integer(n),sortgrid=TRUE)


Env2

```

**Simulating From the Standard Model** 

   This function uses the information contained in the constructed envelope list in order to sample 
 from a model in standard form. The simulation proceeds as follows in order to generate each draw in the required
 number of samples.
 
 1)  A random number between 0 and 1 is generated and is used together with the information in the PLSD vector 
 (from the envelope) in order to identify the part of the grid from which a candidate is to be generated.

 2) For the part of the grid selected, the dimensions are looped through and a candidate component for each dimension
 is generated from a restricted normal using information from the Envelope (in particular, the values for logrt, loglt,  and cbars corresponding to that the part of the grid selected and the dimension sampled)
 
 3) The log-likelihood for the standardized model is evaluated for the generated candidate (note that the 
 log-likelihood here includes the portion of the prior that was shifted to the log-likelihood as part of the standardization procedure)
 
 4) An additional random number is generated and the log of this random number is compared to a log-acceptance rate that is calculated based on the candidate and the LLconst component from the Envelope component selected in order to determine if the candidate should be accepted or rejected 
 
 5) If the candidate was not accepted, the process above is repeated from step 1 until a candidate is accepted


```{r Standard_Simulation}

sim=rnnorm_reg_std(n=as.integer(n),y=as.vector(y),x=as.matrix(x2),mu=as.matrix(mu2,ncol=1),
                   P=as.matrix(P2),alpha=as.vector(alpha),wt=as.vector(wt2),
                   f2=f2,Envelope=Env2,family="binomial",link="logit",as.integer(0))

```

**Undoing the standardization** 


```{r Undo_Standardization}

out=L2Inv%*%L3Inv%*%t(sim$out)

for(i in 1:n){
  out[,i]=out[,i]+mu
}

summary(t(out))
mean(sim$draws)

```

### gaussian families with Independent_Normal_Gamma Prior




## Bounding Distances to Target Densities for Two-Block Gibbs Samplers
