Yaml

---
title: "Chapter 11: Estimating Models with unknown dispersion parameters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chapter 11: Estimating Models with unknown dispersion parameters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning= FALSE,
  message=FALSE
)
```

```{r setup}
library(glmbayes)
```
## 1. Introduction

Earlier chapters developed Bayesian linear and generalized linear models under the assumption
that the dispersion parameter, denoted by \( \phi \) (or equivalently the residual variance
\( \sigma^{2} \)), was known. For Gaussian models this corresponds to fixing the variance of the
error term, and for quasi‑families it corresponds to fixing the scale parameter.

This chapter extends the framework to unknown dispersion parameters, showing how glmbayes
supports:

(1) Priors on the dispersion only  
(2) Joint conjugate Normal–Gamma priors  
(3) Independent Normal–Gamma priors  
      • iid sampling via truncated Gamma priors  
      • two‑block Gibbs sampling  

All gaussian examples use the Dobson plant‑weight data, first introduced in chapter 02.


```{r dobson}
## Annette Dobson (1990) "An Introduction to Generalized Linear Models".
## Page 9: Plant Weight Data.
ctl <- c(4.17, 5.58, 5.18, 6.11, 4.50, 4.61, 5.17, 4.53, 5.33, 5.14)
trt <- c(4.81, 4.17, 4.41, 3.59, 5.87, 3.83, 6.03, 4.89, 4.32, 4.69)
group  <- gl(2, 10, 20, labels = c("Ctl", "Trt"))
weight <- c(ctl, trt)

```


# 2. Exponential-Family Background and Log-Concavity

Many likelihoods used in generalized linear models belong to the exponential family, which can be written in the general weighted form independent observations  
\((y_{1},\ldots,y_{n})\) with weights \((w_{1},\ldots,w_{n})\) can be written as

\[
f(y \mid \theta,\phi,w)
= \exp\!\left\{
  \sum_{i=1}^{n}
    w_{i}
    \left[
      \frac{y_{i}\,\theta_{i} - b(\theta_{i})}{a(\phi)}
      + c(y_{i},\phi)
    \right]
\right\}.
\]

Here:

- \( \theta_{i} \) is the canonical (natural) parameter  
- \( b(\theta_{i}) \) is the cumulant function  
- \( a(\phi) \) is a dispersion or scale parameter  
- \( c(y_{i},\phi) \) collects terms not involving \( \theta_{i} \)  
- \( w_{i} \) is the observation weight (typically \( w_{i} \ge 0 \))
When observations carry nonnegative weights \( w_i \), the weighted log-likelihood becomes

\[
\ell(\beta,\phi)
= \sum_{i=1}^{n}
  w_i \left[
    \frac{y_i\theta_i - b(\theta_i)}{a(\phi)}
    + c(y_i,\phi)
  \right].
\]

This is the form used by both `glm` and the Bayesian functions `glmb`, `lmb`, `rglmb`, and `rlmb`.

---

## 2.1 Log-Concavity in the Exponential Family

For many canonical-link GLMs (Gaussian, Poisson, Binomial, Gamma with log link), the following properties hold.

1. **For fixed dispersion \( \phi \)**,  
   the weighted log-likelihood is concave in \( \beta \) because  
   \[
   \theta_i = \theta(\eta_i), \qquad \eta_i = x_i^{T}\beta,
   \]  
   and \( b(\theta) \) is convex.  
   The weights \( w_i \) simply scale each contribution and preserve concavity.

2. **For fixed coefficients \( \beta \)**,  
   the weighted log-likelihood is often concave in the precision  
   \[
   \tau = \frac{1}{\phi},
   \]  
   because \( a(\phi) \) enters as a positive scalar factor and the weights again preserve concavity.

These concavity properties are essential for:

- unique posterior modes  
- valid tangent-envelope construction for accept-reject sampling  
- stable Gibbs updates for dispersion parameters  
- efficient iid sampling in the glmbayes framework  

---

# 2.2 Special Case: Gaussian Linear Regression

A weighted Gaussian linear model with response vector \( y \in \mathbb{R}^{n} \), design matrix \( X \), coefficients \( \beta \), weights \( w_i \ge 0 \), and dispersion \( \phi \) has log-likelihood

\[
\ell(\beta,\phi)
=
-\frac{1}{2\phi}
\sum_{i=1}^{n}
w_i (y_i - x_i^{T}\beta)^{2}
\;-\;
\frac{1}{2}
\sum_{i=1}^{n}
w_i \log(2\pi\phi).
\]

This matches the exponential-family form with

\[
\theta = \mu = X\beta, \qquad
b(\theta) = \tfrac{1}{2}\theta^{2}, \qquad
a(\phi) = \phi, \qquad
c(y,\phi) = -\tfrac{1}{2\phi}y^{2} - \tfrac{1}{2}\log(2\pi\phi).
\]

Weighted least squares in base R is implemented by transforming the data as

\[
y^{\ast} = W^{1/2} y, \qquad
X^{\ast} = W^{1/2} X,
\]

where \( W = \mathrm{diag}(w_1,\ldots,w_n) \).  
The glmbayes functions use the same weighted likelihood, so the theoretical expressions above apply directly.

---

## Log-Concavity Properties in the Weighted Gaussian Case

1. **Concavity in \( \beta \)**  
   For fixed \( \phi \), the negative weighted log-likelihood is  
   \[
   -\ell(\beta \mid \phi)
   =
   \frac{1}{2\phi}
   (y - X\beta)^{T} W (y - X\beta)
   + \text{const},
   \]
   which is strictly convex whenever \( W \) has positive diagonal entries.  
   Therefore the weighted log-likelihood is strictly concave in \( \beta \).

2. **Concavity in precision \( \tau = 1/\phi \)**  
   Rewriting the weighted likelihood in terms of \( \tau \):
   \[
   \ell(\beta,\tau)
   =
   \frac{\tau}{2}
   (y - X\beta)^{T} W (y - X\beta)
   - \frac{1}{2}
   \left( \sum_{i=1}^{n} w_i \right)
   \log \tau
   + \text{const},
   \]
   which is concave in \( \tau \) because the first term is linear in \( \tau \) and the second is concave.

These properties guarantee:

- unique posterior modes for both \( \beta \) and \( \tau \)  
- efficient iid sampling under Normal, Normal-Gamma, and Independent Normal-Gamma priors  
- stable two-block Gibbs sampling when the dispersion is updated separately  

The weighted Gaussian model thus provides the cleanest illustration of how weights enter the exponential-family structure and how they preserve log-concavity.

## 2.3 Special Case: Gamma Regression With a Log Link

The Gamma family with a log link, as implemented in the glm function, provides another important illustration of exponential–family structure and log–concavity. In this model the mean is

\[
\mu_{i} = \exp(\eta_{i}), \qquad \eta_{i} = x_{i}^{T}\beta,
\]

and the response satisfies

\[
Y_{i} \mid \beta,\phi \sim \text{Gamma}\!\left(\text{shape} = \frac{1}{\phi},\ \text{scale} = \phi \mu_{i}\right),
\]

so that \( \mathbb{E}[Y_{i}] = \mu_{i} \) and \( \text{Var}(Y_{i}) = \phi \mu_{i}^{2} \).  
The corresponding canonical parameter is

\[
\theta_{i} = -\frac{1}{\mu_{i}},
\]

and the cumulant function is

\[
b(\theta_{i}) = -\log(-\theta_{i}).
\]

The dispersion parameter \( \phi \) (or equivalently the precision \( v = 1/\phi \)) enters only through the scalar function \( a(\phi) = \phi \), consistent with the exponential–family form.

### Log–Concavity Properties

For the Gamma log–link model, two key concavity results hold:

1. **Concavity in \( \beta \) for fixed dispersion \( \phi \)**  
   Because the log link ensures  
   \[
   \eta_{i} = x_{i}^{T}\beta, \qquad \mu_{i} = \exp(\eta_{i}),
   \]
   and because the Gamma log–likelihood is concave in the canonical parameter, the overall log–likelihood is concave in \( \beta \).  
   This guarantees a unique posterior mode and stable envelope construction for sampling.

2. **Concavity in the precision \( v = 1/\phi \) for fixed \( \beta \)**  
   The log–likelihood contribution for a single observation can be written as
   \[
   \ell_{i}(v)
   = v \log\!\left(\frac{v}{\mu_{i}}\right)
     - \log\Gamma(v)
     + (v - 1)\log y_{i}
     - \frac{v y_{i}}{\mu_{i}},
   \]
   which is concave in \( v \) because the terms involving \( \log\Gamma(v) \) and \( v \log v \) have negative second derivatives.  
   This concavity is essential for constructing tangent–based envelopes for accept–reject sampling.


### Implications for Bayesian GLMs

The Gamma log–link model thus shares the same structural advantages as the Gaussian and Poisson cases:

* concavity in both \( \beta \) and \( v \),
* unique posterior modes,
* stable envelope construction,
* efficient iid sampling for the dispersion parameter,
* compatibility with two–block Gibbs sampling when desired.

These properties make the Gamma log–link model an especially clean example of how exponential–family structure supports fast, reliable posterior simulation in glmbayes.

## 3. Gaussian Model With Unknown Dispersion

For the Gaussian model

\[
y = X\beta + \varepsilon,\qquad
\varepsilon \sim N(0,\phi I_{n}),
\]

the likelihood is

\[
p(y \mid \beta,\phi)
\propto
\phi^{-n/2}\exp\!\left(-\frac{S(\beta)}{2\phi}\right),
\]

where

\[
S(\beta) = (y - X\beta)^{T}(y - X\beta)
\]

is the residual sum of squares.

To support the different prior specification below, we use the Prior_Setup function to set and extract default settings for the needed parameters.


```{r Prior_Setup,results = "hide"}
ps <- Prior_Setup(weight ~ group)
x  <- ps$x
mu <- ps$mu
V  <- ps$Sigma
y <- ps$y
shape    <- ps$shape
rate     <- ps$rate
```


## 3.1 Priors for the Dispersion Parameter Only

A conjugate prior for the precision (inverse dispersion)\( \tau = 1/\phi \) is

\[
\tau \sim \mathrm{Gamma}(a_{0}, b_{0}).
\]

Given \( \beta \), the posterior is then also a gamma distribution

\[
\tau \mid \beta,y
\sim
\mathrm{Gamma}\!\left(
  a_{0} + \frac{n}{2},\;
  b_{0} + \frac{S(\beta)}{2}
\right).
\]

To illustrate this, we estimate the classical glm model and treat the resulting coefficients as fixed. We can then generate conditional Bayesian draws for the dispersion as below. This is essentially the equivalent of the classical function **gamma.dispersion()** function provided by the **MASS** package. 


```{r glm_call_gamma_prior}
## Call to glm
lm.D9 <- lm(weight ~ group, x = TRUE, y = TRUE)

out_rlmb <- rlmb(
    n = 1000,
    y = y, x = x,
    pfamily = dGamma(shape = shape, rate = rate,
                     beta = lm.D9$coefficients)
  )


```


The posterior mean of the dispersion is then:

```{r glm_call_gamma_mean}
  mean(out_rlmb$dispersion)
```

## 3.2 Joint Conjugate Normal–Gamma Prior

A conjugate prior for the Gaussian model is

\[
\beta \mid \tau \sim N(\mu_{0},\, \tau^{-1}\Sigma_{0}),
\qquad
\tau \sim \mathrm{Gamma}(a_{0}, b_{0}).
\]

The defining property of conjugacy is that the **posterior has the same posterio functional form**. In glmbayes, this model can be estimated as below.


```{r Conjugate_prior}
## Conjugate Normal_Gamma Prior

lmb.D9_v2 <- lmb(
  weight ~ group,
  pfamily = dNormal_Gamma(
    ps$mu,
    ps$Sigma / ps$dispersion,
    shape = ps$shape,
    rate  = ps$rate
  )
)
summary(lmb.D9_v2)$dispersion

```

## 3.3 Independent Normal-Gamma Prior

The independent Normal-Gamma prior removes the conjugate coupling between coefficients and dispersion:

\[
\beta \sim N(\mu_{0}, \Sigma_{0}), \qquad
\tau \sim \text{Gamma}(a_{0}, b_{0}),
\]

so that \( \beta \) and \( \tau \) are a priori independent.

### 3.3.1 Independent Normal Gamma Prior in glmbayes

This prior is implemented in glmbayes using the below specification (but where a truncated gamma replaces the full gamma).

```{r Independent_normal_gamma_prior}
lmb.D9_v3 <- lmb(n=10000,
  weight ~ group,
  dIndependent_Normal_Gamma(
    ps$mu,
    ps$Sigma,
    shape = ps$shape,
    rate  = ps$rate,
  max_disp_perc = 0.99,
  disp_lower = NULL,
  disp_upper = NULL
  )
)
summary(lmb.D9_v3)$coefficients


summary(lmb.D9_v3)$dispersion

sd(lmb.D9_v3$dispersion)


```


### 3.3.2 Two-Block Gibbs Sampling using glmbayes

If iid sampling is not desired, or if the user wants to inspect mixing behavior, a simple two-block Gibbs sampler may be used:

1. Sample \( \beta \mid \tau, y \) from a Normal distribution  
2. Sample \( \tau \mid \beta, y \) from a Gamma distribution

Example:


```{r two_block_Gibbs_sampler}

set.seed(180)


dispersion2 <- ps$dispersion

## Burn-in
for (i in 1:1000) {
  out1 <- rlmb(
    n = 1, y = y, x = x,
    pfamily = dNormal(mu = mu, Sigma = V, dispersion = dispersion2)
  )
  out2 <- rlmb(
    n = 1, y = y, x = x,
    pfamily = dGamma(shape = shape, rate = rate,
                     beta = out1$coefficients[1, ])
  )
  dispersion2 <- out2$dispersion
}

## Sampling
beta_out <- matrix(0, nrow = 10000, ncol = 2)
disp_out <- rep(0, 10000)

for (i in 1:10000) {
  out1 <- rlmb(
    n = 1, y = y, x = x,
    pfamily = dNormal(mu = mu, Sigma = V, dispersion = dispersion2)
  )
  out2 <- rlmb(
    n = 1, y = y, x = x,
    pfamily = dGamma(shape = shape, rate = rate,
                     beta = out1$coefficients[1, ])
  )
  beta_out[i, ] <- out1$coefficients[1, 1:2]
  disp_out[i]   <- out2$dispersion
}

colMeans(beta_out)
sd(beta_out[,1])
sd(beta_out[,2])
mean(disp_out)
sd(disp_out)
```

### 3.3.3 Comparison of the Two Samplers

Under default settings, both the truncated‑Gamma iid sampler and the two‑block Gibbs sampler produce nearly identical posterior summaries for the regression coefficients and the dispersion parameter.  

To make this explicit, the table below reports posterior means and standard deviations from:

* the iid sampler using the independent Normal–Gamma prior (fit via `lmb.D9_v3`), and  
* the two‑block Gibbs sampler constructed from alternating Normal and Gamma updates.

These values illustrate that both samplers target essentially the same posterior distribution (although the former has a truncated prior), with only minor Monte Carlo variation.

| Parameter        | iid Mean | iid SD | Gibbs Mean | Gibbs SD |
|------------------|----------|--------|------------|----------|
| Intercept        | 5.0290   | 0.2323 | 5.0304     | 0.2322   |
| groupTrt         | -0.3648  | 0.3296 | -0.3655    | 0.3298   |
| Dispersion       | 0.5372   | 0.1775 | 0.5415     | 0.1984   |

The difference in the iid and Gibbs SD for the dispersion is largely due to the difference between a truncated and an unrestricted gamma prior.

However, the iid sampler has two clear advantages:

* **Independent draws**: no autocorrelation and no loss of effective sample size  
* **No tuning or burn‑in**: each draw is a direct sample from the posterior

The Gibbs sampler mixes well for this model class, but it inevitably shows mild autocorrelation and convergence diagnostics are beyond the scope of this vignette.  

For most users, the iid sampler is easier to work with and safer as no convergence diagnostics are needed.

For readers who want the full mathematical details of the iid sampler, see **Chapter A7**, which documents
the complete envelope construction and accept‑reject algorithm for the independent Normal–Gamma prior.



## 4. Gamma regression models with unknown dispersion parameter



```{r Carinsca}
data(carinsca)
carinsca$Merit <- ordered(carinsca$Merit)
carinsca$Class <- factor(carinsca$Class)
options(contrasts=c("contr.treatment","contr.treatment"))
Claims=carinsca$Claims
Insured=carinsca$Insured
Merit=carinsca$Merit
Class=carinsca$Class
Cost=carinsca$Cost
```



```{r Prior_Setup_gamma,results = "hide"}
ps=Prior_Setup(Cost/Claims~Merit+Class,family=Gamma(link="log"),weights=Claims)
mu=ps$mu
V=ps$Sigma
shape    <- ps$shape
rate     <- ps$rate
x  <- ps$x
y  <- ps$y

m<-nrow(x)
p<-ncol(x)

```



## 4.1 Priors for the Dispersion Parameter Only

A prior for the precision (inverse dispersion)\( \tau = 1/\phi \) is

\[
\tau \sim \mathrm{Gamma}(a_{0}, b_{0}).
\]



```{r glm_call_gamma_prior2}
## Call to glm
glm.carinsca <- glm(Cost/Claims~Merit+Class,family=Gamma(link="log"),weights=Claims,x=TRUE)

gamma.dispersion(glm.carinsca)

beta = glm.carinsca$coefficients
 
out2 <- rglmb(
    n = 1000, y = y, x = x,
    family  =Gamma(link="log"),
    pfamily = dGamma(shape = shape, rate = rate,
                     beta = glm.carinsca$coefficients),
weights = Claims
  )

mean(out2$dispersion)

```




## 4.2 Indepdent Normal Gamma Prior 

Just like for the gaussian() regression, we can can give the regression coefficients and the inverse dispersion independent priors.

\[
\beta \sim N(\mu_{0}, \Sigma_{0}), \qquad
\tau \sim \text{Gamma}(a_{0}, b_{0}),
\]

so that \( \beta \) and \( \tau \) are a priori independent.

```{r Block_Gibbs_gamma_Regression}

suppressWarnings(
  suppressMessages(
for(i in 1:1000){

  ## --- Block 1: Regression coefficients ---
  out1 <- rglmb(
    n = 1, y = y, x = x,
    family  = Gamma(link="log"),
    pfamily = dNormal(mu = mu, Sigma = V, dispersion = dispersion2),
    weights = Claims
  )

  ## --- Block 2: Dispersion (quasi-likelihood sampler) ---
  out2 <- rglmb(
    n = 1, y = y, x = x,
    family  = Gamma(link="log"),
    pfamily = dGamma(shape = shape, rate = rate,
                     beta = out1$coefficients[1,]),
    weights = Claims
  )

  ## --- SCALE dispersion for the next beta update ---
  ## Convert quasi  (from out2) to MLE for consistency
  dispersion2 <- out2$dispersion * ((m - p)/m)

}

)
)

## Run 1000 additional iterations and store output
beta_out <- matrix(0, nrow = 1000, ncol = 8)
disp_out <- rep(0, 1000)


suppressWarnings(
  suppressMessages(
    for (i in 1:1000) {
      out1 <- rglmb(
        n = 1, y = y, x = x,
        family = Gamma(link="log"),
        pfamily = dNormal(mu = mu, Sigma = V, dispersion = dispersion2),
        weights = Claims
      )
      out2 <- rglmb(
        n = 1, y = y, x = x,
        family = Gamma(link="log"),
        pfamily = dGamma(shape = shape, rate = rate,
                         beta = out1$coefficients[1,]),
        weights = Claims
      )
      dispersion2 <- out2$dispersion * ((m - p) / m)
      beta_out[i, 1:8] <- out1$coefficients[1, 1:8]
      disp_out[i] <- out2$dispersion * ((m - p) / m)
    }
  )
)

## Review output


beta_mean  <- colMeans(beta_out)
beta_sd    <- apply(beta_out, 2, sd)
beta_tlike <- beta_mean / beta_sd   # analogous to GLM t-values

bayes_coef_table <- data.frame(
  Estimate = beta_mean,
  Std.Error = beta_sd,
  t.like = beta_tlike
)

bayes_coef_table

mean(disp_out)


```




### 4.3 Conditional Sampling for the Dispersion Parameter

Chapter A6 provides a detailed derivation of the conditional posterior for the precision parameter \( v = 1/\phi \) in Gamma regression. The posterior can be written in the form

\[
\ell(v)
= ( \text{shape2} - 1 ) \log v
  - \text{rate1} \, v
  + \text{testfunc}(v)
  + \text{const},
\]

where \( \text{testfunc}(v) \) is a concave remainder term capturing the curvature contributed by the Gamma likelihood.  
This decomposition enables a **tangent–envelope accept–reject sampler**, in which:

1. A Gamma proposal distribution is constructed using the linear part  
   \[
   ( \text{shape2} - 1 ) \log v - \text{rate2} \, v,
   \]
   where \( \text{rate2} = \text{rate1} - g^{*} \) and \( g^{*} \) is the derivative of \( \text{testfunc}(v) \) at the anchor point \( v^{*} \).

2. A proposed value \( v_{\text{prop}} \) is accepted with probability  
   \[
   \exp\!\left[
     \text{testfunc}(v_{\text{prop}})
     - \text{testfunc}(v^{*})
     - g^{*}(v_{\text{prop}} - v^{*})
   \right].
   \]

Because \( \text{testfunc}(v) \) is concave, the tangent line at \( v^{*} \) always lies above it, ensuring a valid rejection sampler.




## 8. Summary

This chapter introduced Bayesian estimation of Gaussian models with unknown dispersion,
including:

• Gamma priors on the precision \( \tau \)  
• Joint Normal–Gamma priors  
• Independent Normal–Gamma priors  
• iid sampling via envelope methods  
• two‑block Gibbs sampling  

The log‑concavity of the Gaussian likelihood in both \( \beta \) and \( \tau \) ensures that all samplers
are stable, efficient, and theoretically well‑behaved.

## A1: Posterior Distribution Details for Conjugate Normal-Gamma prior

### A1.1 Marginal Posterior for \( \tau \)

Integrating out \( \beta \) yields a Gamma posterior **independent of \( \beta \)**:

\[
\tau \mid y
\sim
\mathrm{Gamma}\!\left(
  a_{0} + \frac{n}{2},\;
  b_{0} + \frac{1}{2} S_{\text{marg}}
\right),
\]

where

\[
S_{\text{marg}}
=
(y - X\mu_{0})^{T}
\left(I + X\Sigma_{0}X^{T}\right)^{-1}
(y - X\mu_{0}).
\]

This closed‑form marginal is what makes the Normal–Gamma prior so computationally attractive:  
sampling \( \tau \) requires **only a single Gamma draw**.

### A1.2 Conditional Posterior for \( \beta \)

Given \( \tau \), the posterior for \( \beta \) is Gaussian:

\[
\beta \mid \tau,y \sim N(m_{\text{post}}, V_{\text{post}}),
\]

with

\[
V_{\text{post}}
=
\left(\Sigma_{0}^{-1} + \tau X^{T}X\right)^{-1},
\qquad
m_{\text{post}}
=
V_{\text{post}}
\left(\Sigma_{0}^{-1}\mu_{0} + \tau X^{T}y\right).
\]

### A1.3 Sampling the Posterior

Because the Normal–Gamma prior is *conjugate* to the Gaussian likelihood, the **joint posterior factorizes into two standard distributions**:

\[
p(\tau,\beta \mid y)
= p(\tau \mid y)\, p(\beta \mid \tau, y).
\]

This structure means **no MCMC is required**.  
Posterior draws are obtained by **independent Monte Carlo sampling**, not by a Gibbs chain.

---

#### Step 1: Draw the marginal posterior for \( \tau \)

The dispersion (precision) parameter has a closed‑form marginal posterior:

\[
\tau \mid y \sim 
\mathrm{Gamma}\!\left(
  a_{0} + \frac{n}{2},\;
  b_{0} + \frac{1}{2} S(\hat\beta_{\text{GLS}})
\right),
\]

where

\[
S(\hat\beta_{\text{GLS}})
= (y - X\hat\beta_{\text{GLS}})^{T}(y - X\hat\beta_{\text{GLS}})
\]

and \( \hat\beta_{\text{GLS}} \) is the generalized least‑squares estimator under the prior precision.

Each draw of \( \tau \) is **i.i.d.** from this Gamma distribution.

---

#### Step 2: Draw \( \beta \) conditionally on each sampled \( \tau \)

Given a sampled value \( \tau^{(s)} \), the regression coefficients have a Normal posterior:

\[
\beta \mid \tau^{(s)}, y
\sim
N\!\left(m_{\text{post}}^{(s)},\, V_{\text{post}}^{(s)}\right),
\]

with

\[
V_{\text{post}}^{(s)}
=
\left(\Sigma_{0}^{-1} + \tau^{(s)} X^{T}X\right)^{-1},
\]

\[
m_{\text{post}}^{(s)}
=
V_{\text{post}}^{(s)}
\left(\Sigma_{0}^{-1}\mu_{0} + \tau^{(s)} X^{T}y\right).
\]

Each draw of \( \beta \) is **conditionally independent** given its corresponding \( \tau^{(s)} \).

---

### Result: Pure i.i.d. Monte Carlo Sampling

The conjugate posterior allows:

- **direct sampling** of \( \tau \) from its marginal Gamma distribution  
- **direct sampling** of \( \beta \) from its conditional Normal distribution  

No Markov chain is constructed, and **no Gibbs or Metropolis steps** are involved.  
All posterior samples are **independent**, making the conjugate Normal–Gamma model one of the simplest Bayesian regression frameworks for simulation.

## A2. Detailed sampling procedured for the Independent Normal-Gamma prior

### A2.1 iid Sampling Under Truncated Gamma Priors

When iid sampling is requested, glmbayes uses a truncated Gamma prior for the precision parameter \( \tau \).  
The truncation is introduced for numerical stability and does not change the basic Normal-Gamma structure.

The iid sampler for this prior family is based on the envelope accept-reject method described in:

**Chapter A7 — Accept-Reject Sampling for Gaussian Regression Models with Independent Normal-Gamma Priors**

That vignette explains the full simulation procedure, including:

* how the dispersion parameter is truncated,
* how the envelope is constructed across dispersion values,
* how the mixture of truncated normal proposals for \( \beta \) is formed,
* how the Gamma proposal for \( \tau \) is calibrated.

The key point for users is that the posterior remains log-concave, so the accept-reject sampler produces independent draws for \( (\beta, \tau) \).  
There is no Markov chain, no burn-in, and no autocorrelation.

