---
title: "Chapter 08: Case Study — Building Custom GLM Kernels (ex_glmbayes)"
author: "Kjell Nygren"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chapter 08: Case Study — Building Custom GLM Kernels (ex_glmbayes)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

Chapters 02–07 describe the infrastructure of `nmathopencl`: the shim layer,
the `nmath` library, the kernel format, and the runner API. This chapter
shows how to put all of it together by walking through a concrete downstream
use-case: the `ex_glmbayes` example built into the package.

`ex_glmbayes` implements GPU-accelerated evaluation of the log-posterior
and gradient (`f2`/`f3`) for Bayesian GLMs (Gaussian, Gamma, binomial with
logit/probit/cloglog links, and Poisson). It is the type of component a
package author would build on top of `nmathopencl` to accelerate their own
likelihood computations. The name `ex_glmbayes` signals that this is an
example, not a core part of the library; it lives entirely in
`ex_glmbayes_*`-prefixed files and within the `ex_glmbayes` C++ namespace.

## Step 1 — Identify the nmath functions your kernel needs

For each GLM family, determine which nmath functions the GPU kernel must
call:

| Kernel | nmath function(s) called | Link function |
|--------|--------------------------|---------------|
| `f2_f3_binomial_logit.cl` | `dbinom_raw` | logit |
| `f2_f3_binomial_probit.cl` | `dbinom_raw`, `pnorm5`, `dnorm4` | probit |
| `f2_f3_binomial_cloglog.cl` | `dbinom_raw` | cloglog |
| `f2_f3_gamma.cl` | `dgamma` | log |
| `f2_f3_gaussian.cl` | `dnorm4` | identity/log |
| `f2_f3_poisson.cl` | `lgamma` (OpenCL built-in) | log |

For `f2_f3_poisson`, `lgamma` is an OpenCL built-in (part of the OpenCL 1.1
math specification) so no nmath files are needed at all. For the others,
consult the `@all_depends_nmath` tag of each kernel file to get the full
transitive dependency list.

## Step 2 — Extract the minimal nmath subset

The `@all_depends_nmath` tag in each kernel file gives the complete
transitive closure. Taking the union across the five non-Poisson kernels
yields 17 files:

```
Rmath, nmath, refactored,
chebyshev, cospi, fmax2, gammalims, lgammacor,
gamma, lgamma,
stirlerr_cycle_free, pgamma_utils, stirlerr_cycle_dependent, stirlerr, bd0,
dpois, dgamma, dnorm, pnorm, dbinom
```

These 20 files (17 nmath functions + `Rmath`, `nmath`, `refactored` as
infrastructure) were copied to `inst/cl/ex_glmbayes_nmath/`. A package
developer would create an analogous subdirectory for their own kernel set.

The `ex_glmbayes_nmath/` subdirectory is loaded at runtime with:

```{r, eval = FALSE}
nmath_src <- load_kernel_library("ex_glmbayes_nmath", package = "nmathopencl")
```

## Step 3 — Write the kernel files

Each kernel file in `inst/cl/ex_glmbayes_src/` starts with the dependency
metadata block followed by the kernel body:

```c
// @library_deps: nmath
// @calls_nmath: dbinom_raw
// @depends_nmath: dbinom
// @all_depends_nmath_count: 17
// @all_depends_nmath: Rmath, nmath, refactored, chebyshev, cospi, fmax2,
//   gammalims, lgammacor, gamma, lgamma, stirlerr_cycle_free, pgamma_utils,
//   stirlerr_cycle_dependent, stirlerr, bd0, dbinom
// @calls_opencl_builtin: (none)

#pragma OPENCL EXTENSION cl_khr_fp64 : enable
#pragma OPENCL EXTENSION cl_khr_printf : enable

#define MAX_L2 64   // upper bound on l2

__kernel void f2_f3_binomial_logit(
    __global const double* X,      // design matrix,  size = l1 x l2, col-major
    __global const double* B,      // grid,            size = m1 x l2
    __global const double* mu,     // prior mean,      length = l2
    __global const double* P,      // prior precision, size = l2 x l2
    __global const double* alpha,  // offset,          length = l1
    __global const double* y,      // response,        length = l1
    __global const double* wt,     // weights,         length = l1
    __global double*       qf,     // out: neg log-posterior, length = m1
    __global double*       grad,   // out: gradient,   size = m1 x l2
    const int l1,
    const int l2,
    const int m1
) {
    int j = get_global_id(0);
    if (j >= m1) return;

    // prior quadratic form: 0.5 * (B_j - mu)' P (B_j - mu)
    double tmp[MAX_L2];
    for (int k = 0; k < l2; ++k) {
        double acc = 0.0;
        for (int l = 0; l < l2; ++l)
            acc += P[k*l2 + l] * (B[j*l2 + l] - mu[l]);
        tmp[k] = acc;
    }
    double res_acc = 0.0;
    for (int k = 0; k < l2; ++k)
        res_acc += 0.5 * (B[j*l2 + k] - mu[k]) * tmp[k];

    // gradient: prior part
    double g_loc[MAX_L2];
    for (int k = 0; k < l2; ++k) g_loc[k] = tmp[k];

    // data likelihood: logistic link + dbinom_raw
    for (int i = 0; i < l1; ++i) {
        double dot = -alpha[i];
        for (int k = 0; k < l2; ++k)
            dot -= X[k*l1 + i] * B[j*l2 + k];
        double e = exp(dot <= 0 ? dot : -dot);
        double p = dot <= 0 ? 1.0/(1.0+e) : e/(1.0+e);
        double q = 1.0 - p;
        res_acc += -dbinom_raw(y[i], wt[i], p, q, 1);
        double resid = p < 0.5 ? p*wt[i] - y[i]*wt[i]
                                : (1-y[i])*wt[i] - q*wt[i];
        for (int k = 0; k < l2; ++k)
            g_loc[k] += X[k*l1 + i] * resid;
    }

    qf[j] = res_acc;
    for (int k = 0; k < l2; ++k)
        grad[k * m1 + j] = g_loc[k];
}
```

Key design choices:

- `MAX_L2 64` — a compile-time upper bound on the number of predictors,
  used to size local arrays (OpenCL C does not support variable-length
  arrays in all implementations).
- Column-major layout for `X` (matches R's memory layout for matrices
  passed from R to C without transposition).
- A single work-item per grid point `j`; all `l1` observations are
  processed sequentially within that work-item. This is appropriate when
  `m1` (number of grid points) is much larger than `l1` (number of
  observations), as is typical for envelope sampling.

## Step 4 — Write the kernel runner (C++)

The kernel runner is the low-level C++ function that manages the OpenCL
lifecycle for this specific kernel. It lives in
`ex_glmbayes_kernel_runners.cpp` inside the `ex_glmbayes::opencl` namespace:

```cpp
// ex_glmbayes_kernel_runners.cpp (abbreviated)
namespace ex_glmbayes {
namespace opencl {

void f2_f3_kernel_runner(
    const std::string& kernel_source,
    const char*        kernel_name,
    int l1, int l2, int m1,
    const std::vector<double>& X_flat,
    const std::vector<double>& B_flat,
    const std::vector<double>& mu_flat,
    const std::vector<double>& P_flat,
    const std::vector<double>& alpha_flat,
    const std::vector<double>& y_flat,
    const std::vector<double>& wt_flat,
    std::vector<double>& qf_flat,
    std::vector<double>& grad_flat
) {
    // Select platform and device
    // Create context, command queue, program, kernel
    // Create input buffers (CL_MEM_READ_ONLY) and output buffers (CL_MEM_WRITE_ONLY)
    // Set kernel arguments
    // Enqueue kernel with global_size = m1
    // Read back qf_flat and grad_flat
    // Release all resources
}

}} // namespace ex_glmbayes::opencl
```

The runner uses the error-handling utilities from `openclPort.h`
(`opencl_make_context_error`, `opencl_status_name`) to produce informative
exceptions on failure.

## Step 5 — Write the kernel wrapper (C++)

The kernel wrapper is the R-facing `[[Rcpp::export]]` function. It
validates and converts R inputs, assembles the program string, selects the
kernel name based on the family/link, calls the runner, and returns results
as R objects. It lives in `ex_glmbayes_kernel_wrappers.cpp`:

```cpp
// ex_glmbayes_kernel_wrappers.cpp (abbreviated)
namespace ex_glmbayes {
namespace opencl {

// [[Rcpp::export]]
Rcpp::List f2_f3_opencl(
    std::string family, std::string link,
    Rcpp::NumericMatrix b,  Rcpp::NumericVector y,
    Rcpp::NumericMatrix x,  Rcpp::NumericMatrix mu,
    Rcpp::NumericMatrix P,  Rcpp::NumericVector alpha,
    Rcpp::NumericVector wt, int progbar)
{
    int l1 = x.nrow(), l2 = x.ncol(), m1 = b.ncol();

    // 1. Flatten R matrices into contiguous vectors
    auto X_flat  = openclPort::flattenMatrix(x);
    auto B_flat  = openclPort::flattenMatrix(b);
    // ... (all inputs)

    // 2. Select kernel name
    std::string kernel_name;
    if (family == "binomial" && link == "logit")
        kernel_name = "f2_f3_binomial_logit";
    else if (family == "binomial" && link == "probit")
        kernel_name = "f2_f3_binomial_probit";
    // ... (all families)

    // 3. Assemble program source
    std::string nmath_src = openclPort::load_kernel_library(
        "ex_glmbayes_nmath", "nmathopencl");
    std::string kernel_src = openclPort::load_kernel_source(
        "ex_glmbayes_src/" + kernel_name + ".cl", "nmathopencl");
    std::string program_source = nmath_src + "\n" + kernel_src;

    // 4. Allocate output buffers
    std::vector<double> qf_flat(m1);
    std::vector<double> grad_flat(static_cast<size_t>(m1) * l2);

    // 5. Run the kernel
    f2_f3_kernel_runner(program_source, kernel_name.c_str(),
                        l1, l2, m1,
                        X_flat, B_flat, mu_flat, P_flat,
                        alpha_flat, y_flat, wt_flat,
                        qf_flat, grad_flat);

    // 6. Wrap outputs as R objects and return
    Rcpp::NumericVector qf(qf_flat.begin(), qf_flat.end());
    // ... (reshape grad_flat into a matrix)
    return Rcpp::List::create(Rcpp::Named("qf") = qf, ...);
}

}} // namespace ex_glmbayes::opencl
```

## Step 6 — Write the Rcpp export wrappers (C++ and R)

The `[[Rcpp::export]]` annotation on `f2_f3_opencl` causes
`Rcpp::compileAttributes()` to generate:

- A `.Call` entry in `RcppExports.cpp`
- An R wrapper in `RcppExports.R`

The R-side convenience function in `ex_glmbayes.R` adds a `has_opencl()`
guard and CPU fallback:

```{r, eval = FALSE}
f2_f3_opencl_R <- function(family, link, b, y, x, mu, P, alpha, wt,
                             use_opencl = TRUE) {
  if (use_opencl && has_opencl()) {
    f2_f3_opencl(family, link, b, y, x, mu, P, alpha, wt, progbar = 0L)
  } else {
    f2_f3_non_opencl(family, link, b, y, x, mu, P, alpha, wt)
  }
}
```

## Step 7 — Implement the CPU fallback

The CPU fallback uses the same mathematical logic as the kernel but
executes in C++ using standard R math calls. The relevant code lives in:

- `ex_glmbayes_famfuncs_binomial.cpp` — binomial family functions
- `ex_glmbayes_famfuncs_poisson.cpp` — Poisson family
- `ex_glmbayes_famfuncs_Gamma.cpp` — Gamma family
- `ex_glmbayes_famfuncs_gaussian.cpp` — Gaussian family
- `ex_glmbayes_EnvelopeEval.cpp` — orchestrates the CPU path
- `ex_glmbayes_EnvelopeSize.cpp` — memory allocation for the envelope

These files are grouped under the `ex_glmbayes::fam` and `ex_glmbayes::env`
sub-namespaces. They use R's standard `<Rmath.h>` functions (the same
mathematical functions as the GPU kernels, but via the C runtime rather
than OpenCL).

## File inventory

The complete set of files contributing to the `ex_glmbayes` example:

### OpenCL source (`inst/cl/`)

| Directory | Files |
|-----------|-------|
| `ex_glmbayes_nmath/` | 20 files: the minimal nmath subset |
| `ex_glmbayes_src/` | 6 files: one `f2_f3_*.cl` kernel per GLM family |
| `ex_glmbayes_draft_src/` | Draft versions (work in progress) |

### C++ source (`src/`)

| File | Namespace | Role |
|------|-----------|------|
| `ex_glmbayes_kernel_runners.cpp` | `ex_glmbayes::opencl` | Low-level OpenCL runner for the `f2_f3` kernels |
| `ex_glmbayes_kernel_wrappers.cpp` | `ex_glmbayes::opencl` | R-facing wrapper: input validation, program assembly, runner dispatch |
| `ex_glmbayes_export_wrappers.cpp` | (Rcpp export) | `[[Rcpp::export]]` entry points for `EnvelopeSize`, `EnvelopeEval`, `glmb_Standardize_Model` |
| `ex_glmbayes_famfuncs_binomial.cpp` | `ex_glmbayes::fam` | CPU binomial likelihood |
| `ex_glmbayes_famfuncs_poisson.cpp` | `ex_glmbayes::fam` | CPU Poisson likelihood |
| `ex_glmbayes_famfuncs_Gamma.cpp` | `ex_glmbayes::fam` | CPU Gamma likelihood |
| `ex_glmbayes_famfuncs_gaussian.cpp` | `ex_glmbayes::fam` | CPU Gaussian likelihood |
| `ex_glmbayes_EnvelopeEval.cpp` | `ex_glmbayes::env` | CPU envelope evaluation (f2/f3 orchestrator) |
| `ex_glmbayes_EnvelopeSize.cpp` | `ex_glmbayes::env` | Envelope memory sizing |
| `ex_glmbayes_glmb_Standardize_Model.cpp` | `ex_glmbayes::sim` | Design matrix standardization |

### C++ headers (`src/`)

| File | Namespace declared |
|------|--------------------|
| `ex_glmbayes_opencl.h` | `ex_glmbayes::opencl` |
| `ex_glmbayes_famfuncs.h` | `ex_glmbayes::fam` |
| `ex_glmbayes_Envelopefuncs.h` | `ex_glmbayes::env` |
| `ex_glmbayes_simfuncs.h` | `ex_glmbayes::sim` |

### R source (`R/`)

| File | Role |
|------|------|
| `ex_glmbayes.R` | R-facing API: `f2_f3_opencl_R`, `f2_f3_non_opencl` |
| `ex_glmbayes_rcpp_wrappers.R` | Thin Rcpp wrappers for the three exported C++ export functions |

## Adapting this pattern for a new package

To build a similar component in your own package:

1. **Identify your nmath dependencies** — look at the `@all_depends_nmath`
   metadata tags of the nmath functions you call. Copy the listed files
   from `nmathopencl/inst/cl/nmath/` into your package's
   `inst/cl/<yourpkg_nmath>/`.

2. **Write your kernel(s)** — add `@library_deps`, `@calls_nmath`, and
   `@all_depends_nmath` tags at the top of each `.cl` file. Store them in
   `inst/cl/<yourpkg_src>/`.

3. **Write a kernel runner** — use `openclPort::opencl_dbl_scalar_kernel_runner`
   directly if your kernel fits the double-scalar layout, or write a custom
   runner following the same OpenCL lifecycle pattern as `f2_f3_kernel_runner`.
   Use `openclPort::opencl_make_context_error` and `opencl_status_name` for
   error handling.

4. **Write a kernel wrapper** — flatten your R inputs with
   `openclPort::flattenMatrix` / `openclPort::copyVector`, assemble the
   program string with `load_kernel_library()` + `load_kernel_source()`,
   call the runner, and return an R-friendly result.

5. **Write a CPU fallback** — implement the same computation using standard
   R math (`<Rmath.h>`) or base-R functions. Guard the GPU path with
   `has_opencl()`.

6. **Add `LinkingTo: nmathopencl`** to your `DESCRIPTION` file so that
   `openclPort.h` is found at compile time.

The `ex_glmbayes` files are the authoritative reference implementation for
all six steps.
