---
title: "Chapter 02: Adding USE_OPENCL and has_opencl() to Your Package"
author: "Kjell Nygren"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chapter 02: Adding USE_OPENCL and has_opencl() to Your Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

Chapter 01 covers getting OpenCL working for `nmathopencl` itself. This chapter
covers the natural next step for **package developers**: adding `USE_OPENCL` and
`has_opencl()` to a package of your own so that it can call OpenCL kernels
(possibly using the `nmathopencl` kernel library) while still building cleanly
on CRAN and on machines without a GPU SDK.

The two helper functions in this chapter are:

| Function | When to use |
|----------|-------------|
| `use_opencl_configure()` | New package, or package with no existing `src/Makevars` |
| `port_to_opencl_configure()` | Package that already has a committed static `src/Makevars` |

Both produce a `configure` script (Linux/macOS) and a `configure.win` script
(Windows) that generate `src/Makevars` dynamically at install time.

## Why a static `src/Makevars` breaks CRAN

Most Rcpp packages have a static committed `src/Makevars` along the lines of:

```makefile
PKG_CXXFLAGS = $(SHLIB_OPENMP_CXXFLAGS)
PKG_LIBS     = $(SHLIB_OPENMP_CXXFLAGS) $(LAPACK_LIBS) $(BLAS_LIBS) $(FLIBS)
```

If you add OpenCL references directly to this file:

```makefile
PKG_CXXFLAGS = $(SHLIB_OPENMP_CXXFLAGS) -DUSE_OPENCL -I/usr/include
PKG_LIBS     = $(SHLIB_OPENMP_CXXFLAGS) -lOpenCL $(LAPACK_LIBS) $(BLAS_LIBS) $(FLIBS)
```

your package **will fail to compile** on any machine without an OpenCL SDK --
including CRAN's build machines, which have no GPU SDK installed. The build
aborts and no binary is produced.

The fix is a pair of configure scripts that **probe for the SDK at install time**
and generate a CPU-only `Makevars` when no SDK is found. The package always
compiles, and the GPU path is activated only where the SDK is genuinely present.

## The configure → USE_OPENCL → has_opencl() chain

Three entities cooperate to give you CRAN-safe optional GPU acceleration:

```
configure / configure.win          (run by R CMD INSTALL)
  |
  |-- detects CL/cl.h + libOpenCL (+ runtime platform probe on Linux)
  |
  v
src/Makevars / src/Makevars.win    (generated; never committed)
  |
  |-- PKG_CXXFLAGS = ...  [ -DUSE_OPENCL ... ]
  |
  v
#ifdef USE_OPENCL                  (in your C++ source)
  |
  |-- guards all GPU code; package compiles cleanly either way
  |
  v
has_opencl()                       (in your R code)
  |
  `-- calls a compiled-in bool that mirrors the compile-time flag;
      returns TRUE only if -DUSE_OPENCL was set at install time
```

On Linux, the configure script goes one step further: it runs a small C probe
(`clGetPlatformIDs`) to verify that at least one OpenCL platform is actually
registered in `/etc/OpenCL/vendors/`, not just that the ICD loader is
installed. `configure.win` (Windows) relies on header detection alone -- the
GPU driver installs the ICD (`OpenCL.dll`) together with itself.

## Adding `has_opencl()` to your package

### C++ side

Add a thin wrapper that exposes the compile-time flag at runtime. If you are
using `Rcpp::export` attributes (the standard Rcpp workflow), add:

```cpp
// src/opencl_status.cpp
#include <Rcpp.h>

// [[Rcpp::export]]
bool _mypkg_has_opencl_cpp() {
#ifdef USE_OPENCL
  return true;
#else
  return false;
#endif
}
```

`compileAttributes()` (run automatically during `devtools::document()`) will
generate the required SEXP wrapper in `RcppExports.cpp`.

If you prefer a plain `.Call()` without Rcpp attributes, the `nmathopencl`
source in `src/nmathopencl_exports.cpp` shows the equivalent plain-C form.

### R side

```r
# R/opencl_status.R

#' Check whether this package was built with OpenCL support
#'
#' @return Logical scalar: \code{TRUE} if the package was installed from source
#'   with an OpenCL SDK detected by the configure script; \code{FALSE} for
#'   prebuilt CRAN/R-Universe binaries and CPU-only source installs.
#' @export
has_opencl <- function() {
  .Call("_mypkg_has_opencl_cpp", PACKAGE = "mypkg")
}
```

Replace `mypkg` with your package name. The function costs nothing at runtime
(one compiled-in bool comparison) and lets R code branch between GPU and CPU
paths without any dynamic linking to OpenCL.

## Case 1: New package with no existing `src/Makevars`

```{r, eval = FALSE}
# From the root of your package:
use_opencl_configure()
```

This writes `configure` and `configure.win` to the package root, sets
`configure` executable on Unix, and prints a setup checklist. Both scripts
always succeed: when no OpenCL SDK is found they write a CPU-only Makevars.

Add the generated files to `.gitignore` (they are build artefacts):

```
src/Makevars
src/Makevars.win
```

## Case 2: Existing package with a static `src/Makevars`

If your package already has a **committed static** `src/Makevars`, use:

```{r, eval = FALSE}
port_to_opencl_configure()
```

The function:

1. Reads your existing `src/Makevars` and extracts the values of
   `PKG_CPPFLAGS`, `PKG_CXXFLAGS`, `PKG_CFLAGS`, and `PKG_LIBS`.
2. Renames `src/Makevars` → `src/Makevars.in` (the maintained source
   template; commit this file).
3. Generates `configure` (Linux/macOS) and `configure.win` (Windows) that
   read `src/Makevars.in` at install time, run OpenCL detection, and write
   the final `src/Makevars` with the OpenCL flags merged in (or omitted for
   CPU-only).
4. Similarly handles `src/Makevars.win` → `src/Makevars.win.in` if present;
   otherwise copies the generic `configure.win` template.
5. Suggests `.gitignore` entries for the generated `src/Makevars` files.

**After porting, maintain `src/Makevars.in` instead of `src/Makevars`.**
`src/Makevars` is generated at install time and should not be committed.

### What is preserved

All content in `src/Makevars.in` that is not one of the four `PKG_*` key
variables is passed through verbatim (comments, blank lines, and any other
make variables you have defined). The four key variables are rebuilt by the
configure script, incorporating your original base values plus the conditional
OpenCL additions.

For example, if your `src/Makevars.in` contains:

```makefile
# Package uses OpenMP and RcppParallel
PKG_CXXFLAGS = $(SHLIB_OPENMP_CXXFLAGS) -I"$(R_LIBRARY_DIR)/RcppParallel/include"
PKG_LIBS     = $(SHLIB_OPENMP_CXXFLAGS) $(LAPACK_LIBS) $(BLAS_LIBS) $(FLIBS) \
               -L"$(R_LIBRARY_DIR)/RcppParallel/lib" -ltbb
```

The generated `src/Makevars` on an OpenCL-enabled machine will be:

```makefile
# Package uses OpenMP and RcppParallel
PKG_CXXFLAGS = $(SHLIB_OPENMP_CXXFLAGS) -I"$(R_LIBRARY_DIR)/RcppParallel/include" -DUSE_OPENCL -I/usr/include
PKG_LIBS     = -L/usr/lib/x86_64-linux-gnu -lOpenCL $(SHLIB_OPENMP_CXXFLAGS) $(LAPACK_LIBS) $(BLAS_LIBS) $(FLIBS) -L"$(R_LIBRARY_DIR)/RcppParallel/lib" -ltbb
```

And on a CPU-only machine `src/Makevars.in` is copied verbatim as
`src/Makevars`, preserving the original flags exactly.

### Caveats

- **`+=` assignments:** If your `src/Makevars` uses append assignments
  (`PKG_LIBS += -lm`), the function warns and you should review the generated
  configure before use.
- **Generated files:** Do not run `port_to_opencl_configure()` on a
  machine-generated `src/Makevars` (one containing absolute paths or
  `-lOpenCL`). The function checks for these patterns and warns. Run it on the
  static source-controlled file.
- **Existing configure scripts:** If `configure` or `configure.win` already
  exists, the function refuses to overwrite without `overwrite = TRUE`. Users
  who already have configure scripts should integrate the OpenCL detection
  block manually; see
  `system.file("configure-templates", "README.md", package = "nmathopencl")`.

## Guarding OpenCL code in C++

All code that depends on OpenCL headers or the OpenCL runtime must be wrapped
in `#ifdef USE_OPENCL`:

```cpp
#include <Rcpp.h>

#ifdef USE_OPENCL
#include <CL/cl.h>
// ... OpenCL device setup, kernel compilation, dispatch ...
#endif

// [[Rcpp::export]]
Rcpp::NumericVector my_gpu_function(Rcpp::NumericVector x) {
#ifdef USE_OPENCL
  // GPU path
  return run_on_gpu(x);
#else
  // CPU fallback
  return run_on_cpu(x);
#endif
}
```

This is the same pattern used throughout `nmathopencl` and `glmbayes`. The
preprocessor guards ensure the package compiles cleanly in both configurations
from a single codebase.

## Testing the CPU-only path before CRAN submission

Always verify the CPU-only build before submitting to CRAN:

```bash
# Linux / macOS: temporarily disable the configure script
mv configure configure.disabled
R CMD INSTALL --preclean .
Rscript -e "library(mypkg); stopifnot(!has_opencl())"
mv configure.disabled configure

# Restore the GPU-enabled build
R CMD INSTALL --preclean .
```

On Windows, rename `configure.win` similarly. This simulates what CRAN's build
machines experience and will expose any `#ifdef USE_OPENCL` guards you may have
missed.

## DESCRIPTION dependencies

If your package uses `nmathopencl`'s kernel-loading infrastructure or
`openclPort.h`, add the following to `DESCRIPTION`:

```
LinkingTo: nmathopencl, Rcpp
Imports: nmathopencl
```

`LinkingTo` makes `openclPort.h` available at compile time for your C++ code.
`Imports` makes the R-level functions (`load_kernel_library`, etc.) available
at runtime. If you only use `nmathopencl` at the R level (not via `LinkingTo`),
`Imports` alone is sufficient.

## Migration note

`use_opencl_configure()` and `port_to_opencl_configure()` are currently in
`nmathopencl` while `opencltools` completes its initial CRAN review. Once
`opencltools` is available on CRAN, both functions will move there and
`nmathopencl` will re-export them -- the same pattern used for the Tier 4
kernel-authoring tools. The function signatures will not change; no action is
required from downstream package authors.

For the full configure template source, detailed environment-variable
documentation, and the migration plan, see:

```{r, eval = FALSE}
system.file("configure-templates", "README.md", package = "nmathopencl")
```
